# Analyze Lidar Data {#lidar_data}

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(RColorBrewer)
library(scales)
library(ggrepel)
library(viridis)
library(kableExtra)
# spatial
library(sf)
library(stars)
library(lwgeom) 
library(mapview) #Interactive maps
library(leafpop) #map html popup
library(lidR) # lidar data
library(elevatr) # elevation data (DEMs)
library(terra) # raster math
# library(spatstat) #point pattern analysis
```

```{r, warning=F, message=F, results='hide'}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)
```

## Overwrite Data Switch

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
#######################################################
#######################################################
# TURN ON/OFF DATA OVERWRITE
# TURNING THIS ON WILL RE-RUN ALL LIDAR PROCESSING
# WHICH TAKES A LONG TIME
#######################################################
#######################################################
overwrite_it <- TRUE
```

## Download Data From Nat'l Map

The [USGS National Map](https://apps.nationalmap.gov/downloader/) was used to obtain a list of file download links for "Elevation Source Data (3DEP) - Lidar, IfSAR" data available marked as "Lidar Point Cloud (LPC)". This download file list was placed in the `data` folder where the code below utilizes it to download data. The "thumbnail" option in the Nat'l Map was used to determine that the "Fugro" data will suffice to cover the BHEF area. Also, downloaded NAIP imagery and 3dep elevation while had ROI drawn in Nat'l Map.

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
#######################################################
#######################################################
# download lidar data
#######################################################
#######################################################
# open download text file
urls <- read.delim("../data/usgs_lidar_data.txt", header = FALSE) %>% 
  dplyr::rename(url_path = 1) %>% 
  dplyr::filter(grepl("FUGRO", toupper(url_path)) == TRUE) %>%
  dplyr::mutate(
    orig_fname = word(gsub("/", " ", url_path), -1)
    , fname_sans_typ = gsub(".laz", "", orig_fname)
  )


# create parent directory for data
  hey_dir <- "../data/lidar/"
  if(dir.exists(hey_dir)==FALSE){
    dir.create(hey_dir)
  }
#loop through to download lidar data
  for(i in 1:nrow(urls)){
    # set up names
    f_nm <- paste0(hey_dir
      , urls$orig_fname[i]
    )
    options(timeout = 60 * 15)
    ########################
    ## download and unzip
    ########################
    if(file.exists(f_nm)==FALSE){
      # download
      download.file(urls$url_path[i], destfile = f_nm)
    }else{
      print(paste0(f_nm, " file already exists"))
    }
  }
#######################################################
#######################################################
# download elevation data
#######################################################
#######################################################
  # open download text file
  urls <- read.delim("../data/usgs_dem_data.txt", header = FALSE) %>% 
    dplyr::rename(url_path = 1) %>% 
    dplyr::filter(grepl("FUGRO", toupper(url_path)) == TRUE) %>%
    dplyr::mutate(
      orig_fname = word(gsub("/", " ", url_path), -1)
      , fname_sans_typ = gsub(".laz", "", orig_fname)
    )
  
  
  # create parent directory for data
    hey_dir <- "../data/dem/"
    if(dir.exists(hey_dir)==FALSE){
      dir.create(hey_dir)
    }
  #loop through to download dem data
    for(i in 1:nrow(urls)){
      # set up names
      f_nm <- paste0(hey_dir
        , urls$orig_fname[i]
      )
      options(timeout = 60 * 15)
      ########################
      ## download and unzip
      ########################
      if(file.exists(f_nm)==FALSE){
        # download
        download.file(urls$url_path[i], destfile = f_nm)
      }else{
        print(paste0(f_nm, " file already exists"))
      }
    }

```

## Load Vector Data

Spatial data was loaded and cleaned in [prior chapter](#vector_data). 

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
# read
forests_bhnf <- sf::st_read("../data/forests_bhnf.gpkg")
bhef_boundary <- sf::st_read("../data/bhef_boundary.gpkg")
bhef_harvests <- sf::st_read("../data/bhef_harvests.gpkg")
research_plots <- sf::st_read("../data/research_plots.gpkg")
stem_map <- sf::st_read("../data/stem_map.gpkg")
rx_fire <- sf::st_read("../data/rx_fire.gpkg")
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  crowns_group <- sf::st_read("../data/crowns_group.gpkg")
}
if(file.exists("../data/tree_tops.gpkg") == TRUE & overwrite_it == FALSE){
  tree_tops <- sf::st_read("../data/tree_tops.gpkg")
}
if(file.exists("../data/chm.tif") == TRUE & overwrite_it == FALSE){
  chm <- stars::read_stars("../data/chm.tif")
}

```

## Load Elevation Data

The [USGS National Map](https://apps.nationalmap.gov/downloader/) was used to obtain a list of file download links for "Elevation Source Data (3DEP)" data. Using `elevatr::get_elev_raster` to get a digital elevation model (DEM) raster (~6.8m resolution) as a backup.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
#######################################################
#######################################################
# read elevation data
#######################################################
#######################################################
  if(file.exists("../data/bhef_elev.tif") == FALSE){
    #######################################################
    # 1 m dem
    #######################################################
      # tifs <- list.files("../data/dem/", pattern = "\\.tif$", full.names = TRUE)
      # 
      # newgrid <- bhef_boundary %>%
      #   sf::st_buffer(dist = 10000) %>% 
      #   sf::st_bbox() %>%
      #   stars::st_as_stars(dx = 1, dy = 1)
      #   # stars::st_as_stars()
      #   
      # # read files
      # for(i in 1:length(tifs)){
      #   f_nm <- paste0("elev_", i)
      #   assign(f_nm,
      #         stars::read_stars(tifs[i]) %>% 
      #           stars::st_warp(newgrid) %>% 
      #           setNames("elev")
      #           # sf::st_transform(crs = sf::st_crs(bhef_boundary))
      #   )
      # }
      # # mosaic together !!!!!!!!!!!!!!! NEED TO REMOVE HARD-CODING OF OBJECT NAMES
      # elev <- stars::st_mosaic(elev_1, elev_2, elev_3, elev_4)
      # remove(list = ls(pattern = "elev_") )
      # bhef_elev <- elev[sf::st_buffer(bhef_boundary %>% dplyr::select(name), dist = 1000)]
    #######################################################
    # not 1 m dem
    #######################################################
      # z =14 is highest resolution (~6.8m)
      elev <- elevatr::get_elev_raster(bhef_boundary, z = 14)
      # sf::st_crs(elev) == sf::st_crs(bhef_boundary)
      bhef_elev <- stars::st_as_stars(elev)
        # sf::st_transform(crs = sf::st_crs(bhef_boundary))
      # save
    stars::write_stars(bhef_elev, "../data/bhef_elev.tif", append = FALSE)
  }else{
    bhef_elev <- stars::read_stars("../data/bhef_elev.tif")
  }
```

### Elevation plot

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# plot
ggplot() + stars::geom_stars(data = bhef_elev[bhef_boundary]) +
  scale_fill_viridis_c(option = "viridis", alpha = 0.9, na.value = "transparent") +
  geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
  labs(
      title = "BHEF Elevation Map"
      , subtitle = sf::st_crs(bhef_elev)$input
    ) +
  xlab("") +
  ylab("") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8)
    # , panel.grid = element_blank()
    , panel.border = element_blank()
  ) +
  guides(
    fill = guide_legend(title="Elev. (m)")
  )

```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(list = c("urls", "elev", "f_nm", "hey_dir", "tifs", "r", "x"))
gc()
```

## Lidar Data Processing

Use the `lidR` package for manipulating and visualizing point cloud data. See the very [helpful book](https://r-lidar.github.io/lidRbook/index.html) by Roussel, J.R., Goodbody, T.R.H., and Tompalski P. (2021) for more information.

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
# list laz files
  lazs <- sort(list.files("../data/lidar/", pattern = "\\.laz$", full.names = TRUE), decreasing = FALSE)
# change projection of DEM 
  # # creating a new regular grid in a new CRS
  # newgrid <- bhef_boundary %>% 
  #   sf::st_transform(crs = sf::st_crs( lidR::readLAS(lazs[1], select = "xyz") )) %>% 
  #   sf::st_bbox() %>%
  #   stars::st_as_stars()
  # # set up old grid to warp back
  # oldgrid <- bhef_boundary %>% 
  #   sf::st_bbox() %>%
  #   stars::st_as_stars()
  # # warping the old raster to the new grid
  # bhef_elev_reproj <- bhef_elev %>%
  #   stars::st_warp(newgrid)
  
  # bhef_elev_reproj <- bhef_elev %>% stars::st_warp(crs = sf::st_crs( lidR::readLAS(lazs[1], select = "xyz") ))
######################################################
# read laz files
######################################################
  # If several files are read at once the returned LAS object is considered as one LAS file.
  # las <- lidR::readLAS(lazs[2], select = "xyz") %>% # load XYZ only
  las <- lidR::readLAS(lazs, select = "xyz") %>% 
     sf::st_transform(crs = sf::st_crs(bhef_elev))
  # lidR::las_check(las)
  # remove duplicate points
  las <- lidR::filter_duplicates(las)
  # lidR::las_check(las)
  # summary(las$Z)
  # sf::st_crs(las)
  # temp_plot <- plot(las, color = "Z", breaks = "quantile", bg = "white")
  # temp_plot
  # # add_treetops3d(temp_plot, tree_tops)
######################################################
# remove outliers
######################################################
  # Use Statistical Outliers Removal (SOR)
    # k = number of neighbors
    # m = multiplier in : avg distance + m * std deviation
    las <- lidR::classify_noise(las, sor(k = 15, m = 7))
    # plot(las, color = "Classification", bg = "white", size = 3)
  # Remove outliers using filter_poi()
    las <- lidR::filter_poi(las, Classification != LASNOISE)
    # plot(las, color = "Z", breaks = "quantile", bg = "white")
  # #repeat process with Isolated Voxels Filter IVF
  #   las <- lidR::classify_noise(las, ivf(res = 5, n = 6))
  #   # plot(las, color = "Classification", bg = "white", size = 3)
  # # Remove outliers using lidR::filter_poi()
  #   las <- lidR::filter_poi(las, Classification != LASNOISE)
    # plot(las, color = "Z", breaks = "quantile", bg = "white")
# #####################################################
# Ground classification
# !!!!!!!!!!!!!!!!!!!!!!!!!! This is computationally intensive :'(
# #####################################################
# ##################################################
# Progressive Morphological Filter The implementation of PMF algorithm in lidR
# is based on the method described in Zhang et al. (2003)
# ##################################################
# # # b numeric. This is the parameter b in Zhang et al. (2003) (eq. 4 and 5).
# # # dh0 numeric. This is dh0 in Zhang et al. (2003) (eq. 7).
# # # dhmax numeric. This is dhmax in Zhang et al. (2003) (eq. 7).
# # # s numeric. This is s in Zhang et al. (2003) (eq. 7).
# # # max_ws numeric. Maximum window size to be used in filtering ground returns. This
# # # limits the number of windows created.
# # # exp logical. The window size can be increased linearly or exponentially (eq. 4 or 5).
# # lidR::util_makeZhangParam(
# #   b = 2,
# #   dh0 = 0.2,
# #   dhmax = 210,
# #   s = 1.2,
# #   max_ws = 20,
# #   exp = FALSE
# # )
# las <- lidR::classify_ground(las, algorithm = pmf(
#   ws = lidR::util_makeZhangParam()$ws
#   , th = lidR::util_makeZhangParam()$th
#   )
# )
}
```

### Height normalization using DEM 

Height normalization using the DEM loaded above subtacts the height of the ground surface from the point cloud at the resolution of the DEM raster data. The resulting normalized point cloud data classifies ground points at 0 meter in height (Z) and all points  >0 meters in height is assumed to be non-ground surface cover.

```{r, warning=F, message=F, fig.width = 12, fig.height = 6, results='hide'}
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
######################################################
# Height normalization using DEM 
    # could create own DEM with lidar data ... 
    # but will use out of the box product for now
######################################################
  # subtract DEM from lidar returns
    nlas <- las - bhef_elev
  # correct for below ground returns
    nlas@data$Z <- ifelse(
        ceiling(nlas@data$Z) == 0 
        | floor(nlas@data$Z) == 0
        | nlas@data$Z <= 0
      , 0, nlas@data$Z
    )
    # nlas <- filter_poi(nlas, Z >= 0) # remove below ground points
    nlas@data$Z <- ifelse(nlas@data$Z <= 0, 0, nlas@data$Z)
  # update classification
    nlas@data$Classification <- ifelse(nlas@data$Z==0, 2, nlas@data$Classification)
  # filter out top 0.2% heights
    nlas <- filter_poi(nlas, Z <= stats::quantile((nlas@data %>% dplyr::filter(Classification==1))$Z, 0.998))
    # plot(nlas, color = "Z", breaks = "pretty", bg = "white")
    # summary(nlas$Z)
    # table(nlas$Classification)
# ######################################################
# # Height normalization using point cloud interpolation
# # !!!!!! must run ground classification above first
# ######################################################
#   # point cloud normalization using interpolation
#     nlas <- normalize_height(las, knnidw())
#     # plot(nlas, color = "Z", breaks = "pretty", bg = "white")
#     # plot(lidR::filter_ground(nlas), color = "Classification", bg = "white")
#     # summary(nlas$Z)
#     # table(nlas$Classification)
    
######################################################
# descriptive stats
######################################################
kable(nlas@data %>% 
        dplyr::group_by(
          Classification
        ) %>% 
        dplyr::summarise(
          # plots = dplyr::n_distinct(plot)
          points = dplyr::n()
          , min_z = min(Z)
          , max_z = max(Z)
          , mean_z = mean(Z)
          , median_z = median(Z)
          , stdev_z = sd(Z)
        ) %>% 
    dplyr::arrange(Classification) %>% 
    dplyr::mutate(
      Classification = dplyr::case_when(
        Classification == 1 ~ "Surface"
        , Classification == 2 ~ "Ground"
        , TRUE ~ "Other"
      )
    )
  , format = "html" 
  , caption = "Point Cloud Summary Statistics for Return Height (Z)"
  , digits = 1
  , col.names = c(
    "Classification"
    , "points"
    , "min"
    , "max"
    , "mean"
    , "median"
    , "st.dev."
  )
  , align=rep('c', 5)
) %>% 
# kable_classic() %>%
add_header_above(c(" " = 2, "Point Return Height (m)" = 5)) %>% 
kable_material(c("striped", "hover")) %>%
# column_spec(., 2, width = "20em") %>% 
kable_styling(font_size = 11) 
    
# ggplot() + geom_histogram(data = (nlas@data %>% dplyr::filter(Classification == 1)), aes(Z), binwidth = 1)

}
```


### Canopy Height model

Points-to-raster algorithm with a resolution of 1 meter. For each pixel of the output raster the function attributes the height of the highest point found in the pixel from the point cloud.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
######################################################
# Canopy Height model
######################################################
  # Points-to-raster algorithm with a resolution of 1 meter
    chm <- lidR::rasterize_canopy(
      nlas
      , res = 1
      #  for each pixel of the output raster the function attributes the height of the highest point found
      , p2r(subcircle = 0.0
            # , na.fill = tin()
            , na.fill = knnidw(
              k = 10
              , p = 2
              , rmax = 5
            )
        )
      , pkg = "terra"
    ) 
    
    # smooth chm pixels with median value in 3x3 matrix
    kernel <- matrix(1,3,3)
    chm_smooth <- terra::focal(chm, w = kernel, fun = median, na.rm = TRUE) %>% 
      stars::st_as_stars()
      # stars::st_warp(crs = sf::st_crs(bhef_boundary)) 
    # %>% 
    #   sf::st_transform(crs = sf::st_crs(bhef_boundary))
    # non smoothed
    chm <- chm %>% 
      stars::st_as_stars()
      # stars::st_warp(crs = sf::st_crs(bhef_boundary))
}
```

#### CHM plot

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
  # plot
  ggplot() +
    stars::geom_stars(data = chm[rx_fire %>% dplyr::slice_head(n = 1)]) +
    # geom_sf(data = sf::st_crop(bhef_boundary, sf::st_bbox(chm_smooth)), alpha = 0, lwd = 0) +
    geom_sf(data = rx_fire %>% dplyr::slice_head(n = 1), alpha = 0, lwd = 1, color = "orangered") +
    scale_fill_viridis_c(option = "mako", alpha = 0.9, na.value = "transparent") +
    labs(
      title = "BHEF Canopy Height Model"
      , subtitle = paste0("sample Rx fire unit (", sf::st_crs(chm)$input, ")")
    ) +
    xlab("") +
    ylab("") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(
      # legend.position = "bottom"
      # , legend.direction = "horizontal"
      axis.text = element_text(size = 8)
      # , panel.grid = element_blank()
      , panel.border = element_blank()
    ) +
    guides(
      fill = guide_legend(title="Hgt. (m)")
    )
```

### Individual Tree Detection (ITD)

Local maximum filtering (LMF) with variable window size. Points below 2 m will equate to a window size of 3 m, while points above 20 meters equate to a window size of 5 m. Anything between 2 and 20 meter will have a non-linear relationship.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
######################################################
# Individual Tree Detection (ITD)
######################################################
# local maximum filtering (LMF) with variable window size
  # points below 2 m will equate to a window size of 3 m, 
    # while points above 20 meters equate to a window size of 5 m. 
    # Anything between 2 and 20 meter will have a non-linear relationship
  # define variable window function
  ws_fn <- function(x) {
      y <- 2.6 * (-(exp(-0.08*(x-2)) - 1)) + 3
      y[x < 2] <- 3
      y[x > 20] <- 5
      return(y)
  }
  # ITD on CHM
  tree_tops <- lidR::locate_trees(chm_smooth, lmf(ws = ws_fn)) %>% 
    # create classes based on Steel et al. 2021
    dplyr::mutate(
      tree_class = dplyr::case_when(
        Z > 8 ~ 3 # canopy
        , Z >= 2 ~ 2 # subcanopy
        , Z >= 1 ~ 1 # understory
        , TRUE ~ 0 # ground
      )
    )
}
```

#### Tree Tops plot

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
  # plot
  ggplot() +
    stars::geom_stars(data = chm[rx_fire %>% dplyr::slice_head(n = 1)]) +
    geom_sf(data = rx_fire %>% dplyr::slice_head(n = 1), alpha = 0, lwd = 1, color = "orangered") +
    geom_sf(data = sf::st_intersection(tree_tops, rx_fire %>% dplyr::slice_head(n = 1))
            , color = viridis::viridis(n=1, direction = -1), alpha = 0.7, shape = "."
    ) +
    scale_fill_viridis_c(option = "mako", alpha = 0.9, na.value = "transparent") +
    labs(
      title = "BHEF Canopy Height Model with Tree Tops Identified (yellow)"
      , subtitle = paste0("sample Rx fire unit (", sf::st_crs(chm)$input, ")")
    ) +
    xlab("") +
    ylab("") +
    theme_bw() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme(
      axis.text = element_text(size = 8)
      # , panel.grid = element_blank()
      , panel.border = element_blank()
    ) +
    guides(
      fill = guide_legend(title="Hgt. (m)")
    )
  
  # hey_plot <- plot(nlas, color = "Z", breaks = "pretty", bg = "white")
  # add_treetops3d(hey_plot, tree_tops <- lidR::locate_trees(chm_smooth, lmf(ws = ws_fn)), col = "black")
  # ?lidR::pixel_metrics()
```

#### Distribution of Tree Top Heights

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
######################################################
# height range of tree tops
######################################################
tree_tops %>% sf::st_set_geometry(NULL) %>% 
  dplyr::mutate(height_ceil = ceiling(Z)) %>% 
  dplyr::group_by(height_ceil) %>% 
  dplyr::summarise(n = n()) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    pct_tot = n / sum(n)
  ) %>% 
  ggplot(.) +
    # geom_col(aes(x = height_ceil, y = n, fill = n)) +
    geom_bar(aes(x = height_ceil, y = pct_tot, fill = n), color = "gray25", stat = "identity") +
    scale_y_continuous(labels = scales::label_percent(accuracy = 1L)) +
    scale_x_continuous(breaks = seq(min(tree_tops$Z), max(tree_tops$Z)+1, 1))+
    scale_fill_viridis_c(alpha = 0.9, option = "mako", direction = -1) +
    labs(
        title = "Distribution of Tree Top Heights (m)"
      ) +
    xlab("Tree Height (m) Bin") +
    ylab("% Tree tops") +
    theme_bw() +
    theme(
      legend.position="none"
    ) 
```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
gc()
```

### Individual Tree Crown (ITC) Segmentation

Segment the canopy height model raster grid as individual tree crowns (ITC). Following the seeds + growing region algorithm described by [Dalponte and Coomes (2016)](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/2041-210X.12575) to create a vector representation of tree crowns. Then using the individual tree crowns to create groupings and summary statistics of height for each interconnected crown group.


```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6, results='hide'}
if(file.exists("../data/crowns_group.gpkg") == TRUE & overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
######################################################
# Individual Tree Segmentation (ITS)
######################################################
# implements an algorithm for tree segmentation based on 
  # Dalponte and Coomes (2016) algorithm (see reference). 
  # This is a seeds + growing region algorithm.
  algo <- lidR::dalponte2016(
    chm = chm 
    , treetops = tree_tops
    # Threshold below which a pixel cannot be a tree
    , th_tree = 2 
    # Pixel is added to a region if its height is greater than the tree height multiplied by this value.
    , th_seed = 0.45 
    # Pixel is added to a region if its height is greater than the current 
      # mean height of the region multiplied by this value.
    , th_cr = 0.55
    # Maximum value of the crown diameter of a detected tree (in pixels)
    , max_cr = 10
    # If treetops contains an attribute with the ID for each tree, the name of this attribute
    , ID = "treeID"
  )
  crowns_st <- algo()
# quick plot
  # plot(crowns_st, col = lidR::pastel.colors(200), main = "Individual Tree Crowns")

# transform crowns stars object to vector data
  crowns_sf <- crowns_st %>% 
    # convert to vector data and merge polygons with identical pixel values
    sf::st_as_sf(as_points = FALSE, merge = TRUE) %>% 
    # transform to same crs as rest of data # transformed las data above instead
    # sf::st_transform(crs = st_crs(bhef_boundary)) %>% 
    dplyr::mutate(
      tree_id = values
      , crown_area = sf::st_area(.)
    ) %>% 
    dplyr::relocate(tree_id, crown_area) %>% 
    dplyr::select(tree_id, crown_area) %>% 
    sf::st_set_precision(1e7) %>% 
    sf::st_make_valid(.) %>% 
    dplyr::filter(sf::st_is_valid(.))
  
# combine crown groups
  crowns_group <- sf::st_cast(
      sf::st_union(crowns_sf)
    , "POLYGON")
  # quick plot
    # plot(crowns_group, col = lidR::pastel.colors(200), main = "Tree Crown Groups")
  
  #create id column
    crowns_group <- crowns_group %>% 
      merge(., data.frame(geo_type = sf::st_geometry_type(crowns_group)), by.x=0, by.y=0, all.x=TRUE) %>% 
      dplyr::mutate(crown_group_id = as.numeric(as.factor(Row.names)))  %>% 
      dplyr::select(-c(geo_type, Row.names)) %>% 
      dplyr::relocate(crown_group_id) %>% 
      sf::st_as_sf(., sf_column_name = "geometry", crs = sf::st_crs(bhef_boundary))

# join back to individual tree crowns and summarize
  crowns_group_sum <- sf::st_intersection(tree_tops, crowns_group) %>% 
    # if need to keep group at tree level... break here
    sf::st_set_geometry(NULL) %>%
    dplyr::group_by(crown_group_id) %>%
    dplyr::summarise(
      count_trees = dplyr::n_distinct(treeID)
      , min_hgt_m = min(Z, na.rm = TRUE) 
      , max_hgt_m = max(Z, na.rm = TRUE) 
      , mean_hgt_m = mean(Z, na.rm = TRUE) 
      , median_hgt_m = median(Z, na.rm = TRUE) 
      , median_hgt_m = median(Z, na.rm = TRUE)
      , quant10_hgt_m = as.numeric( quantile(Z, probs = .10, na.rm = TRUE) )
      , quant25_hgt_m = as.numeric( quantile(Z, probs = .25, na.rm = TRUE) )
      , quant50_hgt_m = as.numeric( quantile(Z, probs = .50, na.rm = TRUE) )
      , quant75_hgt_m = as.numeric( quantile(Z, probs = .75, na.rm = TRUE) )
      , quant90_hgt_m = as.numeric( quantile(Z, probs = .90, na.rm = TRUE) )
      , count_trees_canopy = sum(ifelse(tree_class == 3, 1, 0), na.rm = TRUE)
      , count_trees_subcanopy = sum(ifelse(tree_class == 2, 1, 0), na.rm = TRUE)
      , count_trees_understory = sum(ifelse(tree_class == 1, 1, 0), na.rm = TRUE)
      , count_trees_ground = sum(ifelse(tree_class == 0, 1, 0), na.rm = TRUE)
    ) %>% 
    dplyr::ungroup() %>% 
    dplyr::arrange(crown_group_id) %>%
    dplyr::mutate(
      crown_group_class = dplyr::case_when(
        count_trees == 1 ~ "Individual Tree"
        , count_trees <= 10 
          & count_trees_canopy/count_trees >= .75 ~  "Small Group - Canopy"
        , count_trees <= 10 
          & count_trees_subcanopy/count_trees >= .75 ~  "Small Group - Subcanopy"
        , count_trees <= 10 ~  "Small Group - Mixed"
        , count_trees <= 40 
          & count_trees_canopy/count_trees >= .75 ~  "Medium Group - Canopy"
        , count_trees <= 40 
          & count_trees_subcanopy/count_trees >= .75 ~  "Medium Group - Subcanopy"
        , count_trees <= 40 ~  "Medium Group - Mixed"
        , count_trees_canopy/count_trees >= .75 ~ "Continuous - Canopy"
        , count_trees_subcanopy/count_trees >= .75 ~ "Continuous - Subcanopy"
        , TRUE ~ "Continuous - Mixed"
      )
    )
#attach summary statistics to spatial crown groups
  crowns_group <- crowns_group %>% 
    dplyr::left_join(crowns_group_sum, by = c("crown_group_id" = "crown_group_id")) %>% 
    dplyr::mutate(
      crown_group_area_ha = as.numeric(sf::st_area(.)) / 10000
      , trees_per_ha = count_trees / crown_group_area_ha
    )
}
```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(list = c("kernel", "algo", "lazs", "ws_fn"))
gc()
```

## Lidar-Derived Crown Groups Summary

Tree classes are based on [Steel et al. 2021](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/1365-2745.13764?casa_token=ob4t10qIRrsAAAAA:qYNA0_uawnsPC4EJ6O5Jz2GCeaA2BXNOUUJBIVFiS2ic-Pm9L5FafLhVtIWg4flYGsEsJgWgMNuPx0w). Height thresholds were used to define canopy (>8 m), subcanopy (2–8 m) and understory (1–2 m). Crown groups are defined by the number of trees in a crown group and the composition of canopy and subcanopy trees in the group:

  * **Individual Tree** : single tree (canopy or subcanopy) without interlocking crown
  
  * **Small Group - Canopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Small Group - Subcanopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Small Group - Mixed** : 2-10 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Medium Group - Canopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Medium Group - Subcanopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Medium Group - Mixed** : 11-40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Continuous - Canopy** : >40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Continuous - Subcanopy** : >40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Continuous - Mixed** : >40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# plot
  ggplot() +
    geom_sf(data = sf::st_intersection(crowns_group, rx_fire %>% dplyr::slice_head(n = 1))
      , aes(fill = crown_group_class)
      , lwd = 0
    ) +
    geom_sf(data = rx_fire %>% dplyr::slice_head(n = 1), alpha = 0, lwd = 1, color = "orangered") +
    scale_fill_viridis_d(option = "turbo", alpha = 0.9) + 
    labs(
      title = "BHEF Crown Groups Classes"
      , subtitle = paste0("sample Rx fire unit (", sf::st_crs(crowns_group)$input, ")")
    ) +
    xlab("") +
    ylab("") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(
      legend.position = "bottom"
      , legend.title = element_blank()
      , axis.text = element_text(size = 8)
      # , panel.grid = element_blank()
      , panel.border = element_blank()
    )
  
```

### Crown Group Summary Statistics {#cg_sum}

```{r, warning=F, message=F, fig.width = 12, fig.height = 6}
# table
kable(crowns_group %>% sf::st_set_geometry(NULL) %>% 
        dplyr::group_by(
          crown_group_class
        ) %>% 
        dplyr::summarise(
          # plots = dplyr::n_distinct(plot)
          crown_groups = dplyr::n()
          , mean_m2 = mean(crown_group_area_ha, na.rm = TRUE) * 10000
          , trees_per_ha = mean(trees_per_ha, na.rm = TRUE)
          , mean_count_trees = mean(count_trees, na.rm = TRUE)
          , mean_count_trees_canopy = mean(count_trees_canopy, na.rm = TRUE)
          , mean_count_trees_subcanopy = mean(count_trees_subcanopy, na.rm = TRUE)
          , mean_hgt_m = mean(mean_hgt_m, na.rm = TRUE)
          # , min_hgt_m = min(min_hgt_m, na.rm = TRUE)
          # , max_hgt_m = max(max_hgt_m, na.rm = TRUE)
        ) %>% 
    dplyr::arrange(desc(mean_count_trees))
  , format = "html" 
  , caption = "Tree Crown Groups Summary Statistics"
  , digits = 1
  , col.names = c(
    "Classification"
    , "# crown groups"
    , "area (m2)"
    , "TPH"
    , "# trees"
    , "# canopy trees"
    , "# subcanopy trees"
    , "height (m)"
    # , "min hgt (m)"
    # , "max hgt (m)"
  )
  , align=rep('c', 5)
) %>% 
# kable_classic() %>%
add_header_above(c(" " = 2, "Mean" = 6)) %>%
kable_material(c("striped", "hover")) %>%
kable_styling(font_size = 9) %>% 
column_spec(1, bold = TRUE, width = "1em")


```

## Map Crowns and Rx Fire Area

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# make map
# different background map types: https://leaflet-extras.github.io/leaflet-providers/preview/
# names(leaflet.providers::providers_loaded()$providers)
mapviewOptions(homebutton = FALSE, basemaps = c("Esri.WorldImagery"))
# crop data
bhef_crowns_group <- sf::st_intersection(
  crowns_group 
    # dplyr::filter(crown_group_area_ha <= 2)
  , rx_fire %>% dplyr::select(unit) %>% dplyr::slice_sample(prop = .5)
) %>% 
sf::st_intersection(., bhef_boundary %>% dplyr::select(name))
# bhef_crowns_group <- sf::st_intersection(
#   crowns_group 
#     # dplyr::filter(crown_group_area_ha <= 2)
#   , sf::st_buffer(bhef_boundary %>% dplyr::select(name), dist = -500)
# )
# map
mapview(bhef_boundary
        , color = "black"
        , lwd = 3
        , alpha.regions = 0
        , label = FALSE
        , legend = FALSE
        , popup = FALSE
) +
mapview(rx_fire
        , color = "orangered"
        , lwd = 2
        , alpha.regions = 0
        , label = FALSE
        , legend = FALSE
        , popup = FALSE
) +
mapview(bhef_crowns_group
        , zcol = "crown_group_class"
        , col.regions = viridis::turbo(n=length(unique(bhef_crowns_group$crown_group_class)))
        , lwd = 0
        , alpha.regions = 0.8
        , label = c("crown_group_class")
        , legend = FALSE
          , popup = popupTable(
              bhef_crowns_group
              , zcol = c(
                "crown_group_class"
                , "count_trees"
                , "min_hgt_m"
                , "max_hgt_m"
                , "mean_hgt_m"
                , "count_trees_canopy"
                , "count_trees_subcanopy"
              )
              , row.numbers = FALSE
              , feature.id = FALSE
            )
) 
```

*Note: crown groups shown for only 50% of Rx Fire units (to enhance rendering performance). Crown group data for entire BHEF is still available.*

## Write Out Data

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6, results='hide'}
if(overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
# save cleaned data for reading to R later
sf::st_write(crowns_group, "../data/crowns_group.gpkg", append = FALSE)
sf::st_write(tree_tops, "../data/tree_tops.gpkg", append = FALSE)
stars::write_stars(chm, "../data/chm.tif", append = FALSE)
stars::write_stars(crowns_st, "../data/crowns_st.tif", append = FALSE)
}
```




```{r, warning=FALSE, message=FALSE, eval=FALSE, echo=FALSE, include=FALSE}
xxx <- chm_smooth %>%
  # create classes based on Steel et al. 2021
  dplyr::mutate(
    veg_class = dplyr::case_when(
      values > 8 ~ 3 # canopy
      , values >= 2 ~ 2 # subcanopy
      , values >= 1 ~ 1 # understory
      , TRUE ~ 0 # ground
    )
  ) %>%
  merge() %>%
  setNames("z") %>%
  st_set_dimensions(names = c("x", "y", "band")) %>%
  # select only the class band
  dplyr::slice(band, 2) %>%
  # convert to vector data and merge polygons with identical values
  sf::st_as_sf(as_points = FALSE, merge = TRUE) %>%
  # transform to same crs as rest of data
  sf::st_transform(crs = sf::st_crs(bhef_boundary))

names(xxx)[1] = "veg_class"

ggplot() +
  geom_sf(data = xxx, aes(fill = as.factor(veg_class)), lwd = 0 ) +
  scale_fill_viridis_d() +
  theme_bw()


```


```{r, warning=F, message=FALSE, eval=FALSE, include=FALSE, echo=FALSE, results='hide'}
## Load NAIP Imagery
# tif = system.file("tif/L7_ETMs.tif", package = "stars")
# x = read_stars(tif)
# x[,,,c(6,5,4,3)]
# r = st_rgb(x[,,,c(6,5,4,3)], 3, use_alpha=TRUE) # now R=6,G=5,B=4,alpha=3
# if (require(ggplot2)) {
#  ggplot() + geom_stars(data = r) + scale_fill_identity()
# }



# load with stars
tifs <- list.files("../data/naip/", pattern = "\\.tif$", full.names = TRUE)

# x <- stars::read_stars(imgs[1])
naip_st <- stars::read_stars(tifs[1])
plot(naip_st %>% dplyr::slice(band, 1:3), axes = TRUE)



naip_rgb <- stars::st_rgb(
  naip_st[,,,c(1, 2, 3)]
  , dimension = 3
  , use_alpha = FALSE
  , maxColorValue = 255L
)
 
# ggplot() +
#   stars::geom_stars(data = naip_rgb) +
#   scale_fill_identity()

```



```{r, warning=F, message=FALSE, eval=FALSE, include=FALSE, echo=FALSE, results='hide'}
# apply itd to smoothed kernel chm?
  methods(class = "stars")


# polygon_list = list(rbind(c(-1, 1), c(-1, -1), c(1, -1), c(1, 1), c(-1, 1)))
  polygon_list = list(rbind(
    # c(x, y, z)
    c(sf::st_bbox(bhef_boundary)[1], sf::st_bbox(bhef_boundary)[4]) # upper left
    , c(sf::st_bbox(bhef_boundary)[1], sf::st_bbox(bhef_boundary)[2]) # lower left
    , c(sf::st_bbox(bhef_boundary)[3], sf::st_bbox(bhef_boundary)[2]) # lower right
    , c(sf::st_bbox(bhef_boundary)[3], sf::st_bbox(bhef_boundary)[4]) # upper right
    , c(sf::st_bbox(bhef_boundary)[1], sf::st_bbox(bhef_boundary)[4]) # upper left
    )
  )
  hey_box <- sf::st_polygon(polygon_list) %>% sf::st_sfc(crs = sf::st_crs(bhef_boundary))
  
  ggplot() + 
    geom_sf(data = hey_box, fill = "transparent", lwd = 2, color = "red") + 
    stars::geom_stars(data = bhef_elev[bhef_boundary]) +
    scale_fill_viridis_c(option = "cividis", alpha = 0.9, na.value = "transparent") +
    theme_bw()
  
  
```
