# Analyze Lidar Data {#stem_map}

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(RColorBrewer)
library(scales)
library(ggrepel)
library(viridis)
library(kableExtra)
# spatial
library(sf)
library(lwgeom) 
library(mapview) #Interactive maps
library(leafpop) #map html popup
library(lidR) # lidar data
library(elevatr) # elevation data (DEMs)
# library(spatstat) #point pattern analysis


```

```{r, warning=F, message=F, results='hide'}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)
```

## Download Data From Nat'l Map

The [USGS National Map](https://apps.nationalmap.gov/downloader/) was used to obtain a list of file download links for "Elevation Source Data (3DEP) - Lidar, IfSAR" data available marked as "Lidar Point Cloud (LPC)". This download file list was placed in the `data` folder where the code below utilizes it to download data. The "thumbnail" option in the Nat'l Map was used to determine that the "Fugro" data will suffice to cover the BHEF area. Also, downloaded NAIP imagery while had ROI drawn in Nat'l Map.

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
#######################################################
#######################################################
# lidar data
#######################################################
#######################################################

# open download text file
urls <- read.delim("../data/usgs_lidar_data.txt", header = FALSE) %>% 
  dplyr::rename(url_path = 1) %>% 
  dplyr::filter(grepl("FUGRO", toupper(url_path)) == TRUE) %>%
  dplyr::mutate(
    orig_fname = word(gsub("/", " ", url_path), -1)
    , fname_sans_typ = gsub(".laz", "", orig_fname)
  )


# create parent directory for data
  hey_dir <- "../data/lidar/"
  if(dir.exists(hey_dir)==FALSE){
    dir.create(hey_dir)
  }
#loop through to download lidar data
  for(i in 1:nrow(urls)){
    # set up names
    f_nm <- paste0(hey_dir
      , urls$orig_fname[i]
    )
    options(timeout = 60 * 15)
    ########################
    ## download and unzip
    ########################
    if(file.exists(f_nm)==FALSE){
      # download
      download.file(urls$url_path[i], destfile = f_nm)
    }else{
      print(paste0(f_nm, " file already exists"))
    }
  }
```

NAIP imagery was downloaded from the [USGS Earth Explorer](https://earthexplorer.usgs.gov/).

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
## Load NAIP Imagery

# load with stars
tifs <- list.files("../data/naip/", pattern = "\\.tif$", full.names = TRUE)

# x <- stars::read_stars(imgs[1])
x <- stars::read_stars(tifs[1])
plot(x %>% dplyr::slice(band, 1), axes = TRUE)

r = stars::st_rgb(x[,,,c(1:3)], use_alpha = FALSE)

# ggplot() + 
#   stars::geom_stars(data = r) +
#   scale_fill_identity()

```

## Load Vector Data

Spatial data was loaded and cleaned in [prior chapter](#vector_data). 

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
# save cleaned data for reading to R later
forests_bhnf <- sf::st_read("../data/forests_bhnf.gpkg")
bhef_boundary <- sf::st_read("../data/bhef_boundary.gpkg")
bhef_harvests <- sf::st_read("../data/bhef_harvests.gpkg")
research_plots <- sf::st_read("../data/research_plots.gpkg")
stem_map <- sf::st_read("../data/stem_map.gpkg")
```

## Load lidar Data

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
# read in elevation data
# z =14 is highest resolution (~6.8m)
if(file.exists("../data/bhef_elev.tif") == FALSE){
  elev <- elevatr::get_elev_raster(bhef_boundary, z = 14)
  # sf::st_crs(elev) == sf::st_crs(bhef_boundary)
  bhef_elev <- stars::st_as_stars(elev)
    # sf::st_transform(crs = sf::st_crs(bhef_boundary))
  # save
  stars::write_stars(bhef_elev, "../data/bhef_elev.tif", append = FALSE)
}else{
  bhef_elev <- stars::read_stars("../data/bhef_elev.tif")
}
plot(bhef_elev[bhef_boundary], axes = TRUE)


# list laz files
lazs <- list.files("../data/lidar/", pattern = "\\.laz$", full.names = TRUE)
# If several files are read at once the returned LAS object is considered as one LAS file.
las <- lidR::readLAS(lazs[1], select = "xyz")  # load XYZ only




```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
gc()
```



## Map Harvests, Research Plots, Stem Map

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# make map
# different background map types: https://leaflet-extras.github.io/leaflet-providers/preview/
# names(leaflet.providers::providers_loaded()$providers)
mapviewOptions(homebutton = FALSE, basemaps = c("Esri"))
# map
mapview(bhef_boundary
        , color = "black"
        , lwd = 3
        , alpha.regions = 0
        , label = FALSE
        , legend = FALSE
        , popup = FALSE
) +
mapview(bhef_harvests_l15
        , zcol = "treatment_type_grp"
        , col.regions = viridis::viridis(n=length(unique(bhef_harvests_l15$treatment_type_grp)))
        , alpha.regions = 0.6
        , label = c("lab")
        , legend = FALSE
          , popup = popupTable(
              bhef_harvests_l15
              , zcol = c(
                "year_id"
                , "treatment_type_grp"
                , "activity_name"
              )
              , row.numbers = FALSE
              , feature.id = FALSE
            )
) +
mapview(research_plots
        , zcol = "trt"
        , col.regions = viridis::mako(n=length(unique(research_plots$trt)), direction = -1)
        , lwd = 2
        , col = "gray90"
        , alpha.regions = 0.8
        , label = c("trt")
        , legend = FALSE
          , popup = popupTable(
              research_plots
              , zcol = c(
                "plot"
                , "trt"
                , "harvest_activity_name_1"
                , "harvest_year_id_1"
              )
              , row.numbers = FALSE
              , feature.id = FALSE
            )
)  +
mapview(stem_map
  , zcol = "unit"
  , col.regions = viridis::turbo(n=length(unique(stem_map$unit)), alpha = 0.8)
    # RColorBrewer::brewer.pal(n = length(unique(stem_map$unit)), name = "RdYlBu")
  , cex = 3.5
  , label = c("unit")
  , legend = FALSE
    , popup = popupTable(
        stem_map
        , zcol = c(
          "unit"
          , "plot"
          , "species"
          , "tag"
          , "heightft"
          , "dbhin"
          , "harvest_activity_name_1"
          , "harvest_year_id_1"
        )
        , row.numbers = FALSE
        , feature.id = FALSE
      )
)

```

*Note, only harvests in last 15 years shown*

## Write Out Data

Make a map of the stem mapped trees.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# save cleaned data for reading to R later
sf::st_write(forests_bhnf, "../data/forests_bhnf.gpkg", append = FALSE)
sf::st_write(bhef_boundary, "../data/bhef_boundary.gpkg", append = FALSE)
sf::st_write(bhef_harvests, "../data/bhef_harvests.gpkg", append = FALSE)
sf::st_write(research_plots, "../data/research_plots.gpkg", append = FALSE)
sf::st_write(stem_map, "../data/stem_map.gpkg", append = FALSE)
```
