# Forest Cover Analysis {#cover_analysis}

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(RColorBrewer)
library(scales)
library(ggrepel)
library(viridis)
library(kableExtra)
# plotting
library(ggpubr)
library(pdftools)
library(qpdf)
library(gridExtra)
library(grid)
# spatial
library(sf)
library(stars)

```

```{r, warning=F, message=F, results='hide'}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)
```

## Set Contour Line Interval

```{r, warning=F, message=F, results='hide', fig.width = 10, fig.height = 6}
#######################################################
#######################################################
# Set Contour Line Interval
#######################################################
#######################################################
contour_int_m <- 5
```



## Read Data

[Vector data](#vector_data) and [forest cover data](#lidar_data) created in prior chapters.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
# read
forests_bhnf <- sf::st_read("../data/forests_bhnf.gpkg")
bhef_boundary <- sf::st_read("../data/bhef_boundary.gpkg")
bhef_harvests <- sf::st_read("../data/bhef_harvests.gpkg")
research_plots <- sf::st_read("../data/research_plots.gpkg")
stem_map <- sf::st_read("../data/stem_map.gpkg")
rx_fire <- sf::st_read("../data/rx_fire.gpkg")
# # forest cov
crowns_group <- sf::st_read("../data/crowns_group.gpkg")
# tree_tops <- sf::st_read("../data/tree_tops.gpkg")
# chm <- stars::read_stars("../data/chm.tif")
# elevation
bhef_elev <- stars::read_stars("../data/bhef_elev.tif")
```

### Data Preparation

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
# create dir for pdf plots
  hey_dir <- "../data/pdf"
  if(dir.exists(hey_dir)==FALSE){
    dir.create(hey_dir)
  }else{ # delete all files if folder exists
    file.remove(list.files(hey_dir, full.names = TRUE))
  }

################################################
################################################
# add colors to the crowns group
################################################
################################################
  all_classes <- c(sort(unique(crowns_group$crown_group_class)), "Non-Tree Cover", "Post-Lidar Harvest")
  col_data <- data.frame(
    crown_group_class = all_classes
    , crown_group_class_color = c(
        viridis::turbo(n = length(unique(crowns_group$crown_group_class)), alpha = 0.5)
        , "#48526B80"
        , "#9E967780"
    )
  ) %>% 
  dplyr::mutate(row_n = dplyr::row_number())
  # set up for plotting
  col_data$crown_group_class <- factor(col_data$crown_group_class, levels=unique(col_data$crown_group_class))
  scale_fill_manual_values_fill <- col_data$crown_group_class_color
  names(scale_fill_manual_values_fill) <- paste(col_data$crown_group_class)

  # attach 
  crowns_group <- crowns_group %>% 
    # dplyr::select(-crown_group_class_color) %>% 
    dplyr::left_join(col_data, by = c("crown_group_class"="crown_group_class"))

################################################
################################################
# create contour data
################################################
################################################
# sequence by 5's to get 5m breaks list for contour
seq_brk <- seq(0, 40000, contour_int_m)
# create contour data
bhef_contour <- bhef_elev %>% 
  stars::st_contour(
    na.rm = TRUE
    , contour_lines = FALSE
    , breaks = # classInt::classIntervals(na.omit(as.vector(x[[1]])))$brks
        seq_brk[
          seq_brk > min(na.omit(as.vector(bhef_elev[[1]]))) - contour_int_m 
          & seq_brk < max(na.omit(as.vector(bhef_elev[[1]]))) + contour_int_m
        ]
  )

```

### Read NAIP Data

[National Agriculture Imagery Program (NAIP)](https://naip-usdaonline.hub.arcgis.com/) imagery for the BHEF was *manually* ;/ downloaded from the [USGS Science Data Explorer](https://ers.cr.usgs.gov/). 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
################################################
################################################
# load naip tif with stars
################################################
################################################
naip_st <- stars::read_stars("../data/naip/bhef_naip_20200809_1m.tif") 

# crop
naip_st <- naip_st %>% 
  sf::st_crop(., sf::st_buffer(sf::st_transform(bhef_boundary, crs = sf::st_crs(naip_st)), dist = 500)) %>% 
  setNames("value") %>% 
  dplyr::mutate(value = ifelse(is.na(value), 0, value)) # black out missing band values

# plot(naip_st %>% dplyr::slice(band, 1:3), axes = TRUE)
# plot(naip_st[,,,3:1], axes = TRUE)

# set rgb image
# band 3 = red, band 2 = green, band 1 = blue
naip_rgb <- stars::st_rgb(
  naip_st[,,,3:1]
  , dimension = 3
  , use_alpha = FALSE
  # , stretch = "histogram"
  , probs = c(0.005, 0.995)
  , stretch = "percent"
)
```

## Post-Lidar Acquisition Timber Harvests

Lidar data was collected in 2017. Spatially identify areas within BHEF where timber harvest (not including TSI) treatments occurred in 2017 or later. 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# find harvests after lidar collection
harvests_post_lidar <- bhef_harvests %>% 
  dplyr::filter(year_id >= 2017) %>% 
  sf::st_intersection(bhef_boundary)
# write data
sf::st_write(harvests_post_lidar, "../data/harvests_post_lidar.gpkg", append = FALSE)

# plot
ggplot() +
  geom_sf(data = harvests_post_lidar
      , aes(fill = treatment_type_grp)
      , lwd = 0
    ) +
  geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
  geom_sf(data = rx_fire, alpha = 0, lwd = 1, color = "orangered") +
  geom_sf_label(data = rx_fire, aes(label = unit)
    , label.size = NA
    , alpha = 0
    , size = 3
  ) +
  geom_sf_label(data = harvests_post_lidar, aes(label = year_id)
    , label.size = NA
    , alpha = 0
    , size = 2.5
  ) +
  scale_fill_viridis_d(option = "viridis", alpha = 0.8) + 
  labs(
    title = "BHEF Post-Lidar Acquisition Timber Harvests"
    , subtitle = paste0("Rx fire units")
    , caption = paste0(
      "*Percent of BHEF Area Harvested Post-Lidar: "
      , scales::percent(
        as.numeric(sum(sf::st_area(harvests_post_lidar))) / as.numeric(sum(sf::st_area(bhef_boundary)))
        , accuracy = 0.1
      )
    )
  ) +
  xlab("") +
  ylab("") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    , legend.title = element_blank()
    , axis.text = element_text(size = 8)
    , panel.border = element_blank()
    , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
    , plot.subtitle = element_text(size=8, face="italic", color="orangered", margin = margin(0,0,0,0))
    , plot.caption = element_text(size=10, face="bold.italic")
  )
# save plot
ggplot2::ggsave(
    filename = paste0("../data/pdf/post_lidar_bhef_tmbrhrvst.pdf")
    , plot = ggplot2::last_plot()
    , width = 8.5
    , height = 11
    , units = "in"
    , dpi = "print"
  )
qpdf::pdf_rotate_pages("../data/pdf/post_lidar_bhef_tmbrhrvst.pdf", angle = 270, output = "../data/pdf/00post_lidar_bhef_tmbrhrvst.pdf")
file.remove("../data/pdf/post_lidar_bhef_tmbrhrvst.pdf")
```


There have been **`r scales::comma(as.numeric(sum(sf::st_area(harvests_post_lidar)))/10000, accuracy = 1)` hectares** in BHEF with timber harvest treatments completed after the lidar acquisition in 2017. This represents **`r scales::percent(as.numeric(sum(sf::st_area(harvests_post_lidar))) / as.numeric(sum(sf::st_area(bhef_boundary))), accuracy = 0.1)`** of the total BHEF area (**`r scales::comma( as.numeric(sf::st_area(bhef_boundary))/10000, accuracy = 1)` hectares**).
      

### Join Harvests to Rx Fire Units

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
rx_fire <- rx_fire %>% 
  dplyr::left_join(
    rx_fire %>% 
      dplyr::select(unit) %>% 
      sf::st_intersection(
        sf::st_union(harvests_post_lidar %>% dplyr::select(nffid))
      ) %>% 
      dplyr::mutate(postlidar_harvest_area_m2 = as.numeric(sf::st_area(.))) %>% 
      sf::st_set_geometry(NULL) %>% 
      dplyr::group_by(unit) %>% 
      dplyr::summarise(postlidar_harvest_area_m2 = sum(postlidar_harvest_area_m2)) %>% 
      dplyr::ungroup() %>% 
      dplyr::select(unit, postlidar_harvest_area_m2)
    , by = c("unit" = "unit")
  ) %>% 
  dplyr::mutate(
    has_postlidar_harvest = ifelse(!is.na(postlidar_harvest_area_m2), 1, 0)
    , postlidar_harvest_area_m2 = ifelse(!is.na(postlidar_harvest_area_m2), postlidar_harvest_area_m2, 0)
    , postlidar_harvest_area_ha = ifelse(!is.na(postlidar_harvest_area_m2), postlidar_harvest_area_m2/10000, 0)
    , area_m2 = as.numeric(sf::st_area(.))
    , area_ha = as.numeric(sf::st_area(.)) / 10000
    , pct_postlidar_harvest_area = postlidar_harvest_area_m2 / area_m2
  )


# plot
ggplot(data = rx_fire) + 
  geom_col(aes(y = reorder(unit, pct_postlidar_harvest_area), x = area_ha)
           , width = 0.7, fill = "navy", alpha = 0.5
  ) +
  geom_col(aes(y = reorder(unit, pct_postlidar_harvest_area), x = postlidar_harvest_area_ha)
           , width = 0.7, fill = "navy", alpha = 0.9
  ) +
  geom_text(
    aes(y = reorder(unit, pct_postlidar_harvest_area)
        , x = 0
        , label = scales::percent(pct_postlidar_harvest_area, accuracy = 1)
      )
    , color = "white"
    , size = 3
    , fontface = "bold"
    , hjust = -0.1
    , vjust = 0.1
  ) +
  labs(
      title = "Percent of Rx Fire Unit Area with Timber Harvest Post-Lidar Acquisition"
      # , subtitle = "within BHEF boundary and bounding box of research plot area"
    ) +
  xlab("Area (ha)") +
  ylab("Rx Fire Unit") +
  scale_x_continuous(labels = scales::comma) +
  theme_bw() +
  theme(
    legend.position = "none"
    , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
    , axis.text.y = element_text(size = 10, face = "bold")
  )
```


## Forest Cover Summary

In [prior chapter](#lidar_data), interlocking tree crowns were grouped together to form distinct tree crown classes. Experimental Forest-wide summary statistics were shown [here](#cg_sum). Explore the canopy cover distributions further.

For reference, tree classes are based on [Steel et al. 2021](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/1365-2745.13764?casa_token=ob4t10qIRrsAAAAA:qYNA0_uawnsPC4EJ6O5Jz2GCeaA2BXNOUUJBIVFiS2ic-Pm9L5FafLhVtIWg4flYGsEsJgWgMNuPx0w). Height thresholds were used to define canopy (>8 m), subcanopy (2–8 m) and understory (1–2 m). Crown groups are defined by the number of trees in a crown group and the composition of canopy and subcanopy trees in the group:

  * **Individual Tree** : single tree >= 2m tall in 2017 (canopy or subcanopy) without interlocking crown
  
  * **Small Group - Canopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Small Group - Subcanopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Small Group - Mixed** : 2-10 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Medium Group - Canopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Medium Group - Subcanopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Medium Group - Mixed** : 11-40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Continuous - Canopy** : >40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Continuous - Subcanopy** : >40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Continuous - Mixed** : >40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# create data for defining crown class
crown_group_class <- c(
  "Individual Tree"
  , "Small Group - Canopy"
  , "Small Group - Subcanopy"
  , "Small Group - Mixed"
  , "Medium Group - Canopy"
  , "Medium Group - Subcanopy"
  , "Medium Group - Mixed"
  , "Continuous - Canopy"
  , "Continuous - Subcanopy"
  , "Continuous - Mixed"
  , "Non-Tree Cover"
  , "Post-Lidar Harvest"
)
description <- c(
  "Single tree >= 2m tall in 2017 (canopy or subcanopy) without interlocking crown"
  , "2-10 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)"
  , "2-10 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)"
  , "2-10 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees"
  , "11-40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)"
  , "11-40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)"
  , "11-40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees"
  , ">40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)"
  , ">40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)"
  , ">40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees"
  , "Any ground cover (vegetation or non-vegetation) less than 2 m in height in 2017"
  , "Timber harvest occured after the BHEF Lidar data acquisition in 2017 (based on FACTS)"
)
temp_dta <- data.frame(
  Classification = crown_group_class
  , Description = description
)
# Export PDF
pdf("../data/pdf/01crown_class_desc.pdf", paper = "USr", height = 8.5, width = 11)
grid::grid.draw(gridExtra::tableGrob(
    temp_dta
    , theme = ttheme_default(base_size = 11, padding = unit(c(8, 8), "mm"))
))
dev.off()  


# table
kable(temp_dta
  , format = "html" 
  , caption = "Crown Class Description"
  , col.names = c(
    "Classification"
    , "Description"
  )
) %>% 
kable_styling(font_size = 11) %>% 
column_spec(1, bold = TRUE, width = "5em") %>% 
column_spec(2, width = "40em")
```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(crown_group_class, description, temp_dta)
gc()
```


## Total Area by Crown Group Class

The total area of the BHEF is approximately **`r scales::comma( as.numeric(sf::st_area(bhef_boundary))/10000, accuracy = 1)` hectares (`r scales::comma( as.numeric(sf::st_area(bhef_boundary))/4047, accuracy = 1)` acres)**. 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# set up data
bhef_area_ha <- sum(as.numeric(sf::st_area(bhef_boundary))/10000)
forestcov_area_ha <- sum(as.numeric(sf::st_area(sf::st_intersection(crowns_group, bhef_boundary)))/10000, na.rm = TRUE)
non_forestcov_area_ha <- bhef_area_ha - forestcov_area_ha
temp_dta <- data.frame(
  crown_group_class = "Non-Tree Cover"
  , area_ha = non_forestcov_area_ha
)
crowns_group_area_sum <- crowns_group %>% 
  sf::st_intersection(bhef_boundary) %>% 
  dplyr::mutate(crown_group_area_ha = as.numeric(sf::st_area(.))/10000) %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::group_by(crown_group_class) %>%
  dplyr::summarise(area_ha = sum(crown_group_area_ha, na.rm = TRUE)) %>% 
  dplyr::ungroup() %>% 
  dplyr::bind_rows(temp_dta) %>% 
  dplyr::mutate(
    pct_area = area_ha / bhef_area_ha
  ) %>% 
  dplyr::arrange(desc(area_ha))
# plot
ggplot(data = crowns_group_area_sum) + 
  geom_col(aes(y = reorder(crown_group_class, area_ha), x = area_ha, fill = area_ha), width = 0.7) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::comma(area_ha, accuracy = 1))
    , color = "black"
    , size = 4
    , fontface = "bold"
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = -0.5
  ) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::percent(pct_area, accuracy = 1))
    , color = "black"
    , size = 3
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = 0.5
  ) +
  labs(
      title = "BHEF Area (ha) by Crown Group Class"
      , subtitle = "*Note: a tree is classified as cover >= 2m tall in 2017 based on lidar acquisition"
    ) +
  xlab("Area (ha)") +
  ylab("") +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_viridis_c(alpha = 0.7, option = "cividis", direction = -1) +
  theme_bw() +
  theme(
    legend.position = "none"
    , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
    , plot.subtitle = element_text(size=8, face="italic", color="gray20", margin = margin(0,0,0,0))
    , axis.text.y = element_text(size = 10, face = "bold")
  )
# save plot
ggplot2::ggsave(
    filename = paste0("../data/pdf/02cover_class_bhef_2017.pdf")
    , plot = ggplot2::last_plot()
    , width = 11
    , height = 8.5
    , units = "in"
    , dpi = "print"
  )


```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(bhef_area_ha, forestcov_area_ha, non_forestcov_area_ha, temp_dta)
gc()
```

## No Harvest Post-Lidar Area by Crown Group Class

The total area of the BHEF which has *not* been harvested after the 2017 lidar acquisition is approximately **`r scales::comma(as.numeric(sum(sf::st_area(sf::st_difference(bhef_boundary, sf::st_union(harvests_post_lidar)))))/10000, accuracy = 1)` hectares (`r scales::comma( as.numeric(sum(sf::st_area(sf::st_difference(bhef_boundary, sf::st_union(harvests_post_lidar)))))/4047, accuracy = 1)` acres)**. 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# set up data
bhef_area_ha <- sum(as.numeric(sf::st_area(sf::st_difference(bhef_boundary, sf::st_union(harvests_post_lidar))))/10000)
forestcov_area_ha <- sum(as.numeric(sf::st_area(sf::st_intersection(crowns_group, sf::st_difference(bhef_boundary, sf::st_union(harvests_post_lidar)))))/10000, na.rm = TRUE)
non_forestcov_area_ha <- bhef_area_ha - forestcov_area_ha
temp_dta <- data.frame(
  crown_group_class = "Non-Tree Cover"
  , area_ha = non_forestcov_area_ha
)
crowns_group_area_sum_woharv <- crowns_group %>% 
  sf::st_intersection(sf::st_difference(bhef_boundary, sf::st_union(harvests_post_lidar))) %>% 
  dplyr::mutate(crown_group_area_ha = as.numeric(sf::st_area(.))/10000) %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::group_by(crown_group_class) %>%
  dplyr::summarise(area_ha = sum(crown_group_area_ha, na.rm = TRUE)) %>% 
  dplyr::ungroup() %>% 
  dplyr::bind_rows(temp_dta) %>% 
  dplyr::mutate(
    pct_area = area_ha / bhef_area_ha
  ) %>% 
  dplyr::arrange(desc(area_ha))
# plot
ggplot(data = crowns_group_area_sum_woharv) + 
  geom_col(aes(y = reorder(crown_group_class, area_ha), x = area_ha, fill = area_ha), width = 0.7) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::comma(area_ha, accuracy = 1))
    , color = "black"
    , size = 4
    , fontface = "bold"
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = -0.5
  ) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::percent(pct_area, accuracy = 1))
    , color = "black"
    , size = 3
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = 0.5
  ) +
  labs(
      title = "No Harvest Post-Lidar BHEF Area (ha) by Crown Group Class"
      , subtitle = "*Note: a tree is classified as cover >= 2m tall in 2017 based on lidar acquisition"
    ) +
  xlab("Area (ha)") +
  ylab("") +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_viridis_c(alpha = 0.7, option = "cividis", direction = -1) +
  theme_bw() +
  theme(
    legend.position = "none"
    , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
    , plot.subtitle = element_text(size=8, face="italic", color="gray20", margin = margin(0,0,0,0))
    , axis.text.y = element_text(size = 10, face = "bold")
  )
# save plot
ggplot2::ggsave(
    filename = paste0("../data/pdf/03cover_class_bhef_noharv.pdf")
    , plot = ggplot2::last_plot()
    , width = 11
    , height = 8.5
    , units = "in"
    , dpi = "print"
  )

```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(bhef_area_ha, forestcov_area_ha, non_forestcov_area_ha, temp_dta)
gc()
```

## Spatial Arrangement by Crown Group Class

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# function to map for each cover class
my_full_area_map <- function(my_data, my_class){
  #generate title
  my_title <- paste0(my_class)
  my_caption <- paste0(
    scales::percent(
      (crowns_group_area_sum %>% dplyr::filter(crown_group_class==my_class))$pct_area
      , accuracy = 1
    )
    , " of total BHEF Area"
  )
  #plot
  (
  ggplot() +
      geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
      geom_sf(data = rx_fire, alpha = 0, lwd = 1, color = "orangered") +
      geom_sf(data = my_data %>% 
                dplyr::filter(crown_group_class==my_class)
          , fill = (col_data %>% dplyr::filter(crown_group_class==my_class))$crown_group_class_color
          , lwd = NA
      ) +
      geom_sf_label(data = rx_fire, aes(label = unit)
        , label.size = NA
        , alpha = 0
        , size = 2
      ) +
      labs(
        title = my_title
        , subtitle = "Spatial Arrangement on BHEF"
        , caption = my_caption
      ) +
      xlab("") +
      ylab("") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme_bw() +
      theme(
        legend.position = "none"
        , legend.direction = "horizontal"
        , legend.title = element_blank()
        , axis.text = element_text(size = 8)
        , panel.border = element_blank()
        , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=10, color="gray20", margin = margin(0,0,0,0))
        , plot.caption = element_text(size=10, face="bold.italic")
      )
    
  )
}

# plot with loop to create a plot for each class
my_classes <- sort(unique(crowns_group$crown_group_class))
for (i in 1:length(my_classes)) {
# for (i in 1:3) {
  print( 
    my_full_area_map(
      my_data = crowns_group %>% sf::st_intersection(bhef_boundary)
      , my_class = my_classes[i]
    ) 
  )
}
```

## Rx Fire Unit Summary

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# summarize area by coverage type
temp_rx_crown <- rx_fire %>% 
  dplyr::select(unit) %>% 
  sf::st_difference(
    sf::st_union(harvests_post_lidar %>% dplyr::select(nffid))
  ) %>%
  sf::st_intersection(
    crowns_group %>% dplyr::select(crown_group_class)
  ) %>% 
  dplyr::mutate(area_m2 = as.numeric(sf::st_area(.))) %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::group_by(unit, crown_group_class) %>% 
  dplyr::summarise(area_m2 = sum(area_m2)) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(unit, crown_group_class, area_m2)
# join to base rx fire data
rx_crown_sum <- rx_fire %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::select(unit, postlidar_harvest_area_m2) %>% 
  dplyr::rename(area_m2 = postlidar_harvest_area_m2) %>% 
  dplyr::mutate(crown_group_class = "Post-Lidar Harvest") %>% 
  dplyr::relocate(names(temp_rx_crown)) %>% 
  dplyr::bind_rows(temp_rx_crown) %>% 
  dplyr::left_join(
    rx_fire %>% 
      sf::st_set_geometry(NULL) %>% 
      dplyr::select(unit, area_m2) %>% 
      dplyr::rename(total_area_m2 = area_m2)
    , by = c("unit" = "unit")
  )
# find non-tree cover
rx_crown_sum <- rx_crown_sum %>% 
  dplyr::bind_rows(
    # find left over area
    rx_crown_sum %>% 
      dplyr::group_by(unit) %>% 
      dplyr::summarise(cov_area_m2 = sum(area_m2), total_area_m2 = max(total_area_m2)) %>% 
      dplyr::ungroup() %>% 
      dplyr::mutate(
        area_m2 = total_area_m2 - cov_area_m2
        , crown_group_class = "Non-Tree Cover"
      ) %>% 
      dplyr::select(names(rx_crown_sum)) %>% 
      dplyr::relocate(names(rx_crown_sum))
  ) %>% 
  dplyr::mutate(
    pct_area = area_m2/total_area_m2
    , area_ha = area_m2/10000
  ) %>% 
  dplyr::arrange(unit, crown_group_class) %>% 
  dplyr::left_join(col_data, by = c("crown_group_class" = "crown_group_class"))

# summarize for plot
rx_crown_sum %>% 
  dplyr::mutate(
    cover = dplyr::case_when(
      crown_group_class %in% c("Non-Tree Cover", "Post-Lidar Harvest") ~ crown_group_class
      , startsWith(crown_group_class, "Continuous") ~ "Continuous Tree Cover"
      , TRUE ~ "Sm/Med Grp. & Indiv. Tree Cover"
    )
    , cover_num = dplyr::case_when(
      crown_group_class == "Non-Tree Cover" ~ 4
      , crown_group_class == "Post-Lidar Harvest" ~ 3
      , startsWith(crown_group_class, "Continuous") ~ 1
      , TRUE ~ 2
    )
  ) %>% 
  dplyr::group_by(unit, cover, cover_num) %>% 
  dplyr::summarise(
    pct_area = sum(pct_area)
    , area_m2 = sum(area_m2)
    , total_area_ha = max(total_area_m2)/10000
  ) %>% 
  dplyr::ungroup() %>% 
# plot
ggplot() + 
  geom_col(aes(
      y = reorder(unit, total_area_ha)
      , x = pct_area
      , fill = reorder(cover, -cover_num)
    )
    , width = 0.7
  ) +
  scale_fill_viridis_d(option = "cividis", alpha = 1, direction = -1) +
  geom_text(
    aes(
      y = reorder(unit, total_area_ha)
      , x = pct_area
      , label = scales::percent(
          ifelse(pct_area < 0.05, NA, pct_area)
        , accuracy = 1)
      , group = reorder(cover, -cover_num)
    )
    , position = position_stack(vjust = 0.5)
    , color = "white"
    , size = 3
    , fontface = "bold"
    # , hjust = -0.1
    # , vjust = 0.1
  ) +
  geom_text(
    aes(
      y = reorder(unit, total_area_ha)
      , x = 1.06
      , label = paste0(scales::comma(
            total_area_ha
          , accuracy = 1)
        , " ha"
      )
    )
    , color = "black"
    , size = 4
    , fontface = "bold"
  ) +
  labs(
      title = "Cover Classification of Rx Fire Unit Area"
      , subtitle = "*Note: a tree is classified as cover >= 2m tall in 2017 based on lidar acquisition"
    ) +
  xlab("% of Area") +
  ylab("Rx Fire Unit") +
  scale_x_continuous(labels = scales::percent) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
    , plot.subtitle = element_text(size=8, face="italic", color="gray20", margin = margin(0,0,0,0))
    , axis.text.y = element_text(size = 10, face = "bold")
    , legend.title = element_blank()
  )
# save plot
ggplot2::ggsave(
    filename = paste0("../data/pdf/04cover_class_by_unit.pdf")
    , plot = ggplot2::last_plot()
    , width = 11
    , height = 8.5
    , units = "in"
    , dpi = "print"
  )
```

```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(temp_rx_crown)
gc()
```

### Rx Fire Unit Cover Class

Create a plot function for cover class by fire unit.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
plot_cover_class_bars <- function(my_unit){
  (
    # plot
    ggplot(data = rx_crown_sum %>% dplyr::filter(unit == my_unit)) + 
      geom_col(
        aes(
          y = reorder(crown_group_class, desc(crown_group_class))
          , x = area_ha
          , fill = crown_group_class
        )
        , width = 0.7
      ) +
      scale_fill_manual(values = scale_fill_manual_values_fill) +
      geom_text(
        aes(
          y = reorder(crown_group_class, desc(crown_group_class))
          , x = area_ha + 0.5
          , label = scales::comma(area_ha, accuracy = 1)
        )
        , color = "black"
        , size = 3
        , fontface = "bold"
        # , position = position_dodge(0.9)
        , hjust = 0.6
        , vjust = -0.3
      ) +
      geom_text(
        aes(
          y = reorder(crown_group_class, desc(crown_group_class))
          , x = area_ha + 0.5
          , label = scales::percent(pct_area, accuracy = 1)
        )
        , color = "black"
        , size = 2.5
        # , position = position_dodge(0.9)
        , hjust = 0.4
        , vjust = 0.7
      ) +
      labs(
          # title = paste0("Rx Fire Unit: ", my_unit)
          subtitle = "Area (ha) by Crown Group Class"
        ) +
      xlab("Area (ha)") +
      ylab("") +
      scale_x_continuous(labels = scales::comma) +
      theme_bw() +
      theme(
        legend.position = "none"
        # , plot.title = element_text(size = 12, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=8, color="gray20", margin = margin(0,0,0,0))
        , axis.text.y = element_text(size = 7, face = "bold")
        , axis.text.x = element_text(size = 7)
        , axis.title.x = element_text(size = 7)
      )
  )
}
# print(plot_cover_class_bars(my_unit = rx_fire$unit[11]))
```

### Rx Fire Unit Cover Map

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
plot_cover_class_map <- function(my_unit){
  (
    #########################################
    #########################################
    # map of unit w/ canopy
    #########################################
    #########################################
    ggplot() +
      stars::geom_stars(
        data = naip_rgb[
          rx_fire %>% dplyr::filter(unit == my_unit) %>% 
          sf::st_buffer(dist = 50) %>% 
          sf::st_transform(crs = sf::st_crs(naip_st)) %>% 
          sf::st_bbox()
        ] 
      ) +
      scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
      geom_sf(data = rx_fire %>%
              dplyr::filter(unit == my_unit) %>%
              sf::st_transform(crs = sf::st_crs(naip_st))
          , alpha = 0
          , lwd = 1.5
          , color = "orangered"
      ) +
      geom_sf(data = crowns_group %>%
                sf::st_crop(
                  rx_fire %>% dplyr::filter(unit == my_unit) %>%
                  sf::st_buffer(dist = 50) %>%
                  sf::st_bbox()
                ) %>%
                sf::st_transform(crs = sf::st_crs(naip_st))
          , lwd = NA
          , aes(fill = crown_group_class_color)
      ) +
      geom_sf(data = harvests_post_lidar %>%
                sf::st_union() %>% 
                sf::st_crop(
                  rx_fire %>% dplyr::filter(unit == my_unit) %>%
                  sf::st_buffer(dist = 50) %>%
                  sf::st_bbox()
                ) %>%
                sf::st_transform(crs = sf::st_crs(naip_st))
          , color = "blue"
          , lwd = 1.5
          , fill = (col_data %>% dplyr::filter(crown_group_class == "Post-Lidar Harvest"))$crown_group_class_color
          # , alpha = 0.4
      ) +
      labs(
        title = paste0("Rx Fire Unit: "
                       , my_unit
                       , "   ("
                       , scales::comma((rx_fire %>% dplyr::filter(unit == my_unit))$area_ha, accuracy = 1)
                       , " ha | "
                       , scales::comma((rx_fire %>% dplyr::filter(unit == my_unit))$area_m2/4047, accuracy = 1)
                       , " ac)"
                )
        , subtitle = ifelse(
          (rx_fire %>% dplyr::filter(unit == my_unit))$has_postlidar_harvest == 1
          , "**This unit has post-lidar timber harvest (blue boundary)"
          , ""
        )
      ) +
      xlab("") +
      ylab("") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme_bw() +
      theme(
        legend.position = "none"
        , legend.direction = "horizontal"
        , legend.title = element_blank()
        , axis.text = element_text(size = 8)
        , panel.border = element_blank()
        , plot.title = element_text(size = 14, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=9, face="bold.italic", color="blue", margin = margin(0,0,0,0))
      )
  )
}
# print(plot_cover_class_map(my_unit = rx_fire$unit[15]))
```

### Rx Fire Unit Contour

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
plot_contour_map <- function(my_unit){
  (
    #########################################
    #########################################
    # map of unit w/ contour
    #########################################
    #########################################
    # plot
    ggplot() +
      stars::geom_stars(data = bhef_elev[
        rx_fire %>% dplyr::filter(unit == my_unit) %>% 
        sf::st_buffer(dist = 50) %>% 
        sf::st_bbox()
      ]) +
      scale_fill_viridis_c(option = "viridis", alpha = 0.9, na.value = "transparent") +
      geom_sf(data = bhef_contour %>% 
        sf::st_crop(
          rx_fire %>% dplyr::filter(unit == my_unit) %>% 
          sf::st_buffer(dist = 50) %>% 
          sf::st_bbox()
        )
        , alpha = 0
        , lwd = 0.5
        , color = "gray30"
        , show.legend = FALSE
      ) +
      geom_sf(data = rx_fire %>%
              dplyr::filter(unit == my_unit)
          , alpha = 0
          , lwd = 1.5
          , color = "orangered"
          , show.legend = FALSE
      ) +
      labs(
        subtitle = paste0("Contour lines are at "
                          , as.character(contour_int_m)
                          , "-m (~"
                          , as.character(round(contour_int_m * 3.281, 1))
                          , "-ft) intervals"
          )
      ) +
      xlab("") +
      ylab("") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        , legend.box.margin = margin(-21,0,2,0)
        , legend.text = element_text(size = 7)
        , legend.title = element_text(size = 7)
        , axis.text.y = element_text(size = 7)
        , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
        , panel.border = element_blank()
        , plot.subtitle = element_text(size=9, face="italic", color="gray20", margin = margin(0,0,0,0))
      ) +
      guides(
        fill = guide_legend(title="Elev. (m)")
      )
  )
}
# print(plot_contour_map(my_unit = rx_fire$unit[10]))
```

### Rx Fire Unit Insert

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
plot_small_map <- function(my_unit){
  (
    ggplot() +
      geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
      geom_sf(data = rx_fire, alpha = 0, lwd = 0.7, color = "orangered") +
      geom_sf(
        data = rx_fire %>% dplyr::filter(unit == my_unit)
        , fill = "black"
        , alpha = 0.2
        , lwd = 1.5
        , color = "orangered3"
      ) +
      geom_sf_label(
        data = rx_fire
        , aes(label = unit)
        , label.size = NA
        , alpha = 0
        , size = 2.5
      ) +
      xlab("") +
      ylab("") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme_bw() +
      theme(
        axis.text = element_text(size = 5)
        , axis.text.x = element_text(angle = 35)
        , panel.border = element_blank()
      )

  )
}
# print(plot_small_map(my_unit = rx_fire$unit[11]))
```

## Rx Fire Info Sheets

Create info sheets for each Rx Fire unit.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# # set up plot legend once
# p_legend <- 
#   ggplot(col_data) +
#     geom_col(
#       aes(
#         x = row_n
#         , y = 1
#         , fill = crown_group_class
#       )
#     ) +
#     scale_fill_manual(values = scale_fill_manual_values_fill) +
#     geom_text(
#       aes(
#         x = row_n
#         , y = 0.5
#         , label = crown_group_class
#       )
#       , size = 2.5
#       , fontface = "bold"
#       , angle = 90
#       , color = "black"
#     ) +
#     theme_nothing()

# loop through all units to create pdf export
for (i in 1:nrow(rx_fire)) {
# for (i in c(1,3)) {
  # plots
  p_class_map <-  plot_cover_class_map(my_unit = rx_fire$unit[i]) 
  p_class_bars <-  plot_cover_class_bars(my_unit = rx_fire$unit[i]) 
  p_contour_map <-  plot_contour_map(my_unit = rx_fire$unit[i]) 
  p_small_map <-  plot_small_map(my_unit = rx_fire$unit[i]) 
  # arrange plots on one page
  info_sht <-
  print(
    ggpubr::ggarrange(
      p_class_map
      , ggpubr::ggarrange(
        p_small_map
        , p_contour_map
        , p_class_bars
        , nrow = 1
        , ncol = 3
        , widths = c(0.5, 1.2, 1)
      )  
      , nrow = 2
      , ncol = 1
      , heights = c(1.7, 1)
      # , align = "hv"
    )
  )
  # export pdf
  ggplot2::ggsave(
    filename = paste0("../data/pdf/unit", sprintf("%02d", as.numeric(rx_fire$unit[i])),".pdf")
    , plot = info_sht
    , width = 11
    , height = 8.5
    , units = "in"
    , dpi = "print"
  )
}

```


## Rx Fire Info Packet

Combine info sheets for each Rx Fire unit into full packet for printing.


```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# combine all pdfs together
pdfs <- sort(list.files("../data/pdf", pattern = "\\.pdf$", full.names = TRUE))

# combine with pdftools
pdftools::pdf_combine(
  pdfs
  , output = paste0(
    "../data/pdf/"
    , "BHEF_RxFire_Plan_"
    , gsub("-", "", Sys.Date())
    , ".pdf"
  )
)


```

