# Forest Cover Analysis {#cover_analysis}

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(RColorBrewer)
library(scales)
library(ggrepel)
library(viridis)
library(kableExtra)
# plotting
library(ggpubr)
library(cowplot)
library(gridExtra)
# spatial
library(sf)
library(stars)
library(lwgeom) 
library(mapview) #Interactive maps
library(leafpop) #map html popup
library(lidR) # lidar data
library(elevatr) # elevation data (DEMs)
library(terra) # raster math
# library(spatstat) #point pattern analysis


```

```{r, warning=F, message=F, results='hide'}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)
```

## Read Data

[Vector data](#vector_data) and [forest cover data](#lidar_data) created in prior chapters.

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
# read
forests_bhnf <- sf::st_read("../data/forests_bhnf.gpkg")
bhef_boundary <- sf::st_read("../data/bhef_boundary.gpkg")
bhef_harvests <- sf::st_read("../data/bhef_harvests.gpkg")
research_plots <- sf::st_read("../data/research_plots.gpkg")
stem_map <- sf::st_read("../data/stem_map.gpkg")
rx_fire <- sf::st_read("../data/rx_fire.gpkg")
# # forest cov
crowns_group <- sf::st_read("../data/crowns_group.gpkg")
# tree_tops <- sf::st_read("../data/tree_tops.gpkg")
# chm <- stars::read_stars("../data/chm.tif")

# add colors to the crowns group
col_data <- data.frame(
  crown_group_class = sort(unique(crowns_group$crown_group_class))
  , crown_group_class_color = viridis::turbo(n = length(unique(crowns_group$crown_group_class)), alpha = 0.5)
)
crowns_group <- crowns_group %>% 
  # dplyr::select(-crown_group_class_color) %>% 
  dplyr::left_join(col_data, by = c("crown_group_class"="crown_group_class"))
```


## Read NAIP Data

[National Agriculture Imagery Program (NAIP)](https://naip-usdaonline.hub.arcgis.com/) imagery for the BHEF was *manually* ;/ downloaded from the [USGS Science Data Explorer](https://ers.cr.usgs.gov/). 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
# load with stars
naip_st <- stars::read_stars("../data/naip/bhef_naip_20200809_1m.tif") 

# crop
naip_st <- naip_st %>% 
  sf::st_crop(., sf::st_buffer(sf::st_transform(bhef_boundary, crs = sf::st_crs(naip_st)), dist = 500))

# plot(naip_st %>% dplyr::slice(band, 1:3), axes = TRUE)
# plot(naip_st[,,,3:1], axes = TRUE)

# set rgb image
# band 3 = red, band 2 = green, band 1 = blue
naip_rgb <- stars::st_rgb(
  naip_st[,,,3:1]
  , dimension = 3
  , use_alpha = FALSE
  # , stretch = "histogram"
  , probs = c(0.005, 0.995)
  , stretch = "percent"
)

ggplot() +
  stars::geom_stars(data = naip_rgb[
    rx_fire %>% dplyr::slice(3) %>% 
    sf::st_buffer(dist = 50) %>% 
    sf::st_transform(crs = sf::st_crs(naip_st)) %>% 
    sf::st_bbox()
  ]) +
  scale_fill_identity() + # !!! don't take this out or RGB plot will kill your computer
  geom_sf(data = rx_fire %>%
          dplyr::slice(3) %>%
          sf::st_transform(crs = sf::st_crs(naip_st))
      , alpha = 0
      , lwd = 1.5
      , color = "orangered"
  ) +
  geom_sf(data = crowns_group %>%
            sf::st_crop(
              rx_fire %>% dplyr::slice(3) %>%
              sf::st_buffer(dist = 50) %>%
              sf::st_bbox()
            ) %>%
            sf::st_transform(crs = sf::st_crs(naip_st))
      , lwd = 0
      , aes(fill = crown_group_class_color)
  ) +
  xlab("") +
  ylab("") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw() +
  theme(
    legend.position = "none"
    , legend.direction = "horizontal"
    , legend.title = element_blank()
    , axis.text = element_text(size = 8)
    , panel.border = element_blank()
  )

```


## Forest Cover Summary

In [prior chapter](#lidar_data), interlocking tree crowns were grouped together to form distinct tree crown classes. Experimental Forest-wide summary statistics were shown [here](#cg_sum). Explore the canopy cover distributions further.

For reference, tree classes are based on [Steel et al. 2021](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/1365-2745.13764?casa_token=ob4t10qIRrsAAAAA:qYNA0_uawnsPC4EJ6O5Jz2GCeaA2BXNOUUJBIVFiS2ic-Pm9L5FafLhVtIWg4flYGsEsJgWgMNuPx0w). Height thresholds were used to define canopy (>8 m), subcanopy (2–8 m) and understory (1–2 m). Crown groups are defined by the number of trees in a crown group and the composition of canopy and subcanopy trees in the group:

  * **Individual Tree** : single tree (canopy or subcanopy) without interlocking crown
  
  * **Small Group - Canopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Small Group - Subcanopy** : 2-10 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Small Group - Mixed** : 2-10 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Medium Group - Canopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Medium Group - Subcanopy** : 11-40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Medium Group - Mixed** : 11-40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
  * **Continuous - Canopy** : >40 trees with interlocking crowns where 75% or more of the trees are canopy (>8 m)
  
  * **Continuous - Subcanopy** : >40 trees with interlocking crowns where 75% or more of the trees are subcanopy (2-8 m)
  
  * **Continuous - Mixed** : >40 trees with interlocking crowns with neither canopy nor subcanopy comprising >75% of trees
  
## Total Area by Crown Group Class

The total area of the BHEF is approximately **`r scales::comma( as.numeric(sf::st_area(bhef_boundary))/10000, accuracy = 1)` hectares (`r scales::comma( as.numeric(sf::st_area(bhef_boundary))/4047, accuracy = 1)` acres)**. 

```{r, warning=F, message=F, fig.width = 10, fig.height = 6, results='hide'}
# set up data
bhef_area_ha <- as.numeric(sf::st_area(bhef_boundary))/10000
forestcov_area_ha <- sum(crowns_group$crown_group_area_ha, na.rm = TRUE)
non_forestcov_area_ha <- bhef_area_ha - forestcov_area_ha
temp_dta <- data.frame(
  crown_group_class = "Non-Tree Cover"
  , area_ha = non_forestcov_area_ha
)
crowns_group_area_sum <- crowns_group %>% 
  sf::st_set_geometry(NULL) %>% 
  dplyr::group_by(crown_group_class) %>%
  dplyr::summarise(area_ha = sum(crown_group_area_ha, na.rm = TRUE)) %>% 
  dplyr::ungroup() %>% 
  dplyr::bind_rows(temp_dta) %>% 
  dplyr::mutate(
    pct_area = area_ha / bhef_area_ha
  ) %>% 
  dplyr::arrange(desc(area_ha))
# plot
ggplot(data = crowns_group_area_sum) + 
  geom_col(aes(y = reorder(crown_group_class, area_ha), x = area_ha, fill = area_ha), width = 0.7) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::comma(area_ha, accuracy = 1))
    , color = "black"
    , size = 4
    , fontface = "bold"
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = -0.5
  ) +
  geom_text(
    aes(y = reorder(crown_group_class, area_ha), x = area_ha + 2, label = scales::percent(pct_area, accuracy = 1))
    , color = "black"
    , size = 3
    # , position = position_dodge(0.9)
    , hjust = -0.1
    , vjust = 0.5
  ) +
  labs(
      title = "BHEF Area (ha) by Crown Group Class"
      # , subtitle = "within BHEF boundary and bounding box of research plot area"
    ) +
  xlab("Area (ha)") +
  ylab("") +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_viridis_c(alpha = 0.7, option = "cividis", direction = -1) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```


```{r, warning=F, message=F, echo=FALSE, include=FALSE}
remove(bhef_area_ha, forestcov_area_ha, non_forestcov_area_ha, temp_dta)
```

## Spatial Area by Crown Group Class

```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# plot
  ggplot() +
    geom_sf(data = crowns_group %>% dplyr::slice_sample(prop = 0.2)
      , aes(fill = crown_group_class)
      , lwd = 0
    ) +
    facet_wrap(~crown_group_class) +
    geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black") +
    scale_fill_viridis_d(option = "turbo", alpha = 0.9) + 
    labs(
      title = "BHEF Crown Groups Classes"
      , subtitle = paste0("sample Rx fire unit (", sf::st_crs(crowns_group)$input, ")")
    ) +
    xlab("") +
    ylab("") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(
      legend.position = "bottom"
      , legend.title = element_blank()
      , axis.text = element_text(size = 8)
      # , panel.grid = element_blank()
      , panel.border = element_blank()
    )
  
```


```{r, warning=F, message=F, fig.width = 10, fig.height = 6}
# summarize mean temp by year, month, site
metoc_site_year_mo <- metoc_hourly_full %>% 
  group_by(site, year, month) %>% 
  summarise(mean_temp = mean(air.temp..K., na.rm = TRUE)
            , cnt = n()) %>% 
  arrange(site, year, month)

# function to create line graph of temp by year for each site
my_line_graph_month <- function(my_data, my_year){
  #generate title
  my_title <- paste0(as.character(my_year), " Average Temperature by Month")
  #plot
  (
  ggplot(my_data %>% filter(year==my_year), aes(x = (month), y = mean_temp, color = site)) +
    geom_line(alpha=0.8, size = 1.3) + 
    geom_point(alpha=0.8, size = 1.3) + 
    scale_x_discrete(limits=month.abb) +
    labs(
        title = my_title
        , subtitle = "Swamp Angel Study Plot and Senator Beck Study Plot"
      ) +
    ylab("Mean Temp. (K)") +
    xlab("") +
    scale_color_brewer(palette = "Dark2") +
    theme_bw() +
    theme(
      legend.position="bottom"
      , legend.box = "horizontal"
      , axis.text.x = element_text(size = 11)
    )
  )
}

# plot with loop to create a plot for each year
my_years <- c(2005:2010)
for (i in 1:length(my_years)) {
  print( my_line_graph_month(metoc_site_year_mo, my_years[i]) )
}
```



```{r, warning=F, message=F, fig.width = 18, fig.height = 8}
# table
kable(crowns_group %>% sf::st_set_geometry(NULL) %>% 
        dplyr::group_by(
          crown_group_class
        ) %>% 
        dplyr::summarise(
          # plots = dplyr::n_distinct(plot)
          crown_groups = dplyr::n()
          , mean_m2 = mean(crown_group_area_ha, na.rm = TRUE) * 10000
          , trees_per_ha = mean(trees_per_ha, na.rm = TRUE)
          , mean_count_trees = mean(count_trees, na.rm = TRUE)
          , mean_count_trees_canopy = mean(count_trees_canopy, na.rm = TRUE)
          , mean_count_trees_subcanopy = mean(count_trees_subcanopy, na.rm = TRUE)
          , mean_hgt_m = mean(mean_hgt_m, na.rm = TRUE)
          # , min_hgt_m = min(min_hgt_m, na.rm = TRUE)
          # , max_hgt_m = max(max_hgt_m, na.rm = TRUE)
        ) %>% 
    dplyr::arrange(desc(mean_count_trees))
  , format = "html" 
  , caption = "Tree Crown Groups Summary Statistics"
  , digits = 1
  , col.names = c(
    "Classification"
    , "# crown groups"
    , "area (m2)"
    , "TPH"
    , "# trees"
    , "# canopy trees"
    , "# subcanopy trees"
    , "height (m)"
    # , "min hgt (m)"
    # , "max hgt (m)"
  )
  , align=rep('c', 5)
) %>% 
# kable_classic() %>%
add_header_above(c(" " = 2, "Mean" = 6)) %>%
kable_material(c("striped", "hover")) %>%
kable_styling(font_size = 9) %>% 
column_spec(1, bold = TRUE, width = "1em")

```
