# Forest Cover Analysis {#cover_analysis}

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  , results='hide'
  , fig.width = 10
  , fig.height = 8
)
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(RColorBrewer)
library(scales)
library(ggrepel)
library(viridis)
library(kableExtra)
library(cowplot)
# plotting
library(mapview)
# spatial
library(sf)
library(stars)
library(terra)
```

```{r, results='hide'}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)
```

## Read Data

[Vector data](#vector_data) and [forest cover data](#lidar_data) created in prior chapters.

```{r read-data}
# read
bhef_boundary <- sf::st_read("../data/bhef_boundary.gpkg")
# bhef_harvests <- sf::st_read("../data/bhef_harvests.gpkg")
# bhef_tsi <- sf::st_read("../data/bhef_tsi.gpkg")
rx_fire <- sf::st_read("../data/rx_fire.gpkg")
# # forest cov
crowns_group <- sf::st_read("../data/crowns_group.gpkg")
crowns_rast <- terra::rast("../data/crowns_st.tif")
# chm <- terra::rast("../data/chm.tif") %>% 
#   terra::aggregate(fact = 5, fun = "mean", na.rm = T)
# elevation
bhef_elev <- terra::rast("../data/bhef_elev.tif") 
  # terra::aggregate(fact = 5, fun = "mean", na.rm = T)
```

## Aspect and HLI{#hli}

Aspect of a slope (the direction or azimuth that a slope faces) strongly influences potential direct incident radiation and temperature. Untransformed, aspect is a poor variable for quantitative analysis, since 1° is adjacent to 360° – the numbers are very different even though the aspect is about the same. The following equation for Heat Load Index (HLI) rescales aspect to a scale of zero to one, with zero being the coolest slope (northeast) and one being the warmest slope (southwest). HLI is a proxy for aspect-driven solar heating.

$$
\textrm{Heat load index} = \frac{1 - \cos(\theta - 45)}{2}
$$
where $\theta$ = aspect in degrees east of north. While this and related equations are useful, they do not consider the steepness of the slope. For example, a 1° south-facing slope would receive the same heat load index as a 30° south-facing slope, even though the latter will be considerably warmer.

HLI calculations have been developed which combine slope, folded aspect (i.e., “southwestness”), and latitude to estimate terrain-driven solar heating at a given location ([McCune & Keon 2002: eq. 2](https://onlinelibrary.wiley.com/doi/10.1111/j.1654-1103.2002.tb02087.x)). To calculate HLI, we used equation 2 of [McCune & Keon 2002: eq. 2](https://onlinelibrary.wiley.com/doi/10.1111/j.1654-1103.2002.tb02087.x). Higher values of HLI are typical on steep southwesterly slopes, while low values are common on steep northeasterly slopes. Moderate values are common in flatter terrain and on northwesterly and southeasterly aspects. 

Aspect must first be "folded" about the north-south line, rescaling 0-360° to 0-180°, such that NE = NW, E = W, etc: 

$$
\textrm{Folded aspect°} = 180° \, - \, | Aspect° – 180° |
$$


```{r hli-calc, results='hide', fig.show='hide'}
# convert elevation raster from terra to stars
my_crs <- 5070 # EPSG:5070 = NAD83/Conus Albers (units = meters)  
elev_stars_temp <- stars::st_as_stars(bhef_elev |> terra::project( paste0("EPSG:",my_crs) )) %>% 
  setNames(c("elev"))
# extract latitude degrees from stars object
lats_temp <- sf::st_coordinates(elev_stars_temp) %>%
  sf::st_as_sf(coords = c("x", "y"), crs = my_crs) %>%
  sf::st_transform(crs = 4326) %>%
  sf::st_coordinates() %>%
  dplyr::as_tibble() %>%
  setNames(c("longitude","latitude")) %>%
  dplyr::select(latitude)
gc()
# hli function
  # This equation returns ln(Rad, MJ · cm–2 · yr–1). It can
  # be returned to an arithmetic scale with the EXP(x)
  # function.
hli_fn <- function(slope_rad, folded_aspect_rad, latitude_rad){
  exp(
    (-1*1.236) +
      1.350*cos(latitude_rad)*cos(slope_rad) +
      (-1*1.376)*cos(folded_aspect_rad)*sin(slope_rad)*sin(latitude_rad) +
      (-1*0.331)*sin(latitude_rad)*sin(slope_rad) +
      0.375*sin(folded_aspect_rad)*sin(slope_rad)
  )
}
# calculate slope and aspect from DEM
topo_temp <- c(
    elev_stars_temp
    , starsExtra::slope(elev_stars_temp)
    , starsExtra::aspect(elev_stars_temp)
  ) %>% 
  setNames(
    c("elev", "slope_deg", "aspect_deg")
  ) %>% 
  dplyr::mutate(
    slope_rad = pracma::deg2rad(as.numeric(slope_deg))
    , aspect_rad = pracma::deg2rad(as.numeric(aspect_deg))
    , folded_aspect_deg = 180 - abs(as.numeric(aspect_deg)-180)
    , folded_aspect_rad = pracma::deg2rad(folded_aspect_deg)
    , latitude_deg = lats_temp$latitude
    , latitude_rad = pracma::deg2rad(as.numeric(latitude_deg))
    , hli = dplyr::case_when(
        hli_fn(slope_rad, folded_aspect_rad, latitude_rad) > 1 ~ 1
        , hli_fn(slope_rad, folded_aspect_rad, latitude_rad) < 0 ~ 0
        , TRUE ~ hli_fn(slope_rad, folded_aspect_rad, latitude_rad)
      )
  )
# convert just HLI back to terra
hli <- terra::rast((topo_temp %>% dplyr::select(hli))[]) %>% 
  terra::project( terra::crs(crowns_rast) ) %>% 
  # resample to align layers
  terra::resample(y=crowns_rast, method = "bilinear") %>% 
  terra::crop(crowns_rast)
  # rename
  names(hli) <- "hli"
plot(hli, main = "HLI")
# does it have the same extent?
terra::ext(hli) == terra::ext(crowns_rast)
if(FALSE){
# smooth HLI with median value in matrix
  # d = 1 = 3x3 # d = 2 = 5x5 # d = 3 = 7x7
  # fw_temp <- terra::focalMat(hli, d = 2, type = "rectangle")
  fw_temp <- matrix(1,5,5)
  hli <- terra::focal(hli, w = fw_temp, fun = "median", na.rm=T)
  plot(hli, main = "HLI smooth")
  # does it have the same extent?
  terra::ext(hli) == terra::ext(crowns_rast)
}
# aggregate to 25m resolution
hli %>% 
  terra::aggregate(fact = 25, fun = "median") %>% 
  plot(main = "HLI aggregate")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Tree Crown Cover

The tree crown raster data represents tree crowns (interconnected crown groups or individual crowns) with minimum tree height of 2 meters was created [here](#itc) using `lidR::dalponte2016`. Values of this raster are tree crown id's. 

```{r crowns-rast, results='hide', fig.show='hide'}
# values
crowns_rast %>% terra::values() %>% summary()
# plot
crowns_rast %>% 
  as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot() +
    geom_raster(mapping=aes(x=x,y=y,fill=f)) +
    scale_fill_viridis_c()
# zoom plot
crowns_rast %>% 
  terra::crop(
    bhef_boundary %>% sf::st_centroid() %>% sf::st_buffer(10)
  ) %>% 
  as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot() +
    geom_raster(mapping=aes(x=x,y=y,fill=as.factor(f))) +
    scale_fill_viridis_d(option="turbo")
# mapview
mv_temp <- crowns_rast %>% 
  terra::crop(
    bhef_boundary %>% sf::st_centroid() %>% sf::st_buffer(500)
  ) %>% 
  stars::st_as_stars()
mapview::mapviewOptions(homebutton = FALSE, basemaps = c("Esri.WorldImagery"))
mapview::mapview(mv_temp, col.regions = "forestgreen", na.color = "transparent", legend = F)
# reclass crowns as 1, else 0
  # reclass matrix
  rcl_temp <- c(
      crowns_rast %>% terra::values() %>% min(na.rm = T)
      , crowns_rast %>% terra::values() %>% max(na.rm = T)
      , 1
    ) %>%  
    matrix(ncol=3, byrow=TRUE)
  # binary raster
  iscrown_rast <- crowns_rast %>% 
    terra::classify(
      rcl = rcl_temp
      , include.lowest = T
      , others = 0
    ) %>% 
    # update to 0/1 because na values are not 0
    terra::subst(from=NA,to=0)
# values
iscrown_rast %>% terra::values() %>% summary()
iscrown_rast %>% terra::res()
# plot
iscrown_rast %>% 
  as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot() +
    geom_raster(mapping=aes(x=x,y=y,fill=as.factor(f))) +
    geom_sf(data = bhef_boundary, fill = NA, color = "black", lwd = 1.1) +
    scale_fill_viridis_d(na.value = "transparent")
# plot zoom
iscrown_rast %>% 
  terra::crop(
    bhef_boundary %>% sf::st_centroid() %>% sf::st_buffer(10)
  ) %>% 
  as.data.frame(xy=T) %>% 
  dplyr::rename(f=3) %>% 
  ggplot() +
    geom_raster(mapping=aes(x=x,y=y,fill=as.factor(f))) +
    scale_fill_viridis_d(na.value = "transparent")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Aggregate and Crop to Rx fire Units

```{r agg-rast-dta}
# select aggregate resolution: 25 = 25x25m grid
agg_res <- 25
# aggreaged and combine layers
layers <- c(
    iscrown_rast %>% 
      terra::aggregate(fact = agg_res, fun = "mean")
    , hli %>% 
      terra::aggregate(fact = agg_res, fun = "median")
  ) %>% 
  setNames(c("crowncov", "hli"))
# crop to rx fire units bounds
layers <- layers %>% 
  terra::mask(
    rx_fire %>% 
      sf::st_union() %>% 
      terra::vect()
  )
# plot
nm_temp <- c("% Canopy Cov.", "HLI")
plt_fn_temp <- function(layer_n) {
  plt <- layers[[layer_n]] %>% 
    as.data.frame(xy=T) %>% 
    dplyr::rename(f=3) %>% 
    ggplot() +
      geom_raster(mapping=aes(x=x,y=y,fill=f)) +
      geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
      geom_sf(data = rx_fire, alpha = 0, lwd = 1, color = "orangered") +
      geom_sf_label(data = rx_fire, aes(label = unit)
        , label.size = NA
        , alpha = 0
        , size = 3
      ) +
      coord_sf(expand = F) +
      scale_fill_viridis_c(option = (8-layer_n), alpha = 0.8) + 
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      labs(
          title = nm_temp[layer_n]
          , subtitle = paste0("Rx fire units")
          , x=""
          , y=""
          , fill = nm_temp[layer_n]
        ) +
      theme_light() +
      theme(
        legend.position = "bottom"
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        , legend.box.margin = margin(-21,0,2,0)
        , legend.text = element_text(size = 7, angle = 25, hjust = 0.7)
        , legend.title = element_text(size = 7)
        , axis.text = element_text(size = 7)
        , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
        , plot.title = element_text(size = 10, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=8, face="italic", color="orangered", margin = margin(0,0,0,0))
        , panel.border = element_blank()
      )
  return(plt)
}
plt_list_temp <- 1:(layers %>% terra::nlyr()) %>% 
  purrr::map(plt_fn_temp)

# combine
cowplot::plot_grid(
  plotlist = plt_list_temp
  , nrow = 1
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Classify and Sample

```{r class-create}
rast_grp_class_fn <- function(layer_n, pct_cut_list = seq(0,1,0.2)) {
  # classify based on percentile
    cuts_temp <- stats::quantile(
      layers[[layer_n]] %>% terra::values()
      , probs = pct_cut_list
      , na.rm = T
    )
  # create classification raster based on percentiles
    rast_temp <- layers[[layer_n]] %>% 
      classify(
        rcl = cuts_temp
        , include.lowest=TRUE
        , brackets=TRUE
      )
  # recode to get rid of bracket labels and set index to 1 instead of 0 based
    rcl_temp <- rast_temp %>% 
      terra::levels() %>% 
      purrr::pluck(1) %>% 
      dplyr::as_tibble() %>% 
      dplyr::mutate(
        old = value
        , new = value+1
      ) %>% 
      dplyr::select(old,new)
  # classify
    rast_temp <- rast_temp %>% 
      terra::classify(rcl=rcl_temp) %>% 
      terra::categories(
        value = rcl_temp %>% 
          dplyr::mutate(category = new) %>% 
          dplyr::select(new,category) %>% 
          dplyr::rename(id=new)
      )
  # return
  return(rast_temp)
}
# create grouped raster stack
layers_grp <- 1:(layers %>% terra::nlyr()) %>% 
  purrr::map(rast_grp_class_fn) %>% 
  terra::rast() %>% 
  setNames(
    names(layers)
  )
# plot
nm_temp <- c("% Canopy Cov.", "HLI")
plt_fn_temp <- function(layer_n) {
  plt <- layers_grp[[layer_n]] %>% 
    as.data.frame(xy=T) %>% 
    dplyr::rename(f=3) %>% 
    dplyr::mutate(
      f = factor(
        f
        , levels = 1:5
        , labels = c("low","med.low","medium","med.high","high")
        , ordered = T
      )
    ) %>% 
    ggplot() +
      geom_raster(mapping=aes(x=x,y=y,fill=f)) +
      geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
      geom_sf(data = rx_fire, alpha = 0, lwd = 1, color = "orangered") +
      geom_sf_label(data = rx_fire, aes(label = unit)
        , label.size = NA
        , alpha = 0
        , size = 3
      ) +
      coord_sf(expand = F) +
      scale_fill_viridis_d(option = "turbo", alpha = 0.8) + 
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      labs(
          title = nm_temp[layer_n]
          , subtitle = paste0("Rx fire units")
          , x=""
          , y=""
          , fill = nm_temp[layer_n]
        ) +
      theme_light() +
      theme(
        legend.position = "bottom"
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        , legend.box.margin = margin(-21,0,2,0)
        , legend.text = element_text(size = 7)
        , legend.title = element_text(size = 7)
        , axis.text = element_text(size = 7)
        , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
        , plot.title = element_text(size = 10, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=8, face="italic", color="orangered", margin = margin(0,0,0,0))
        , panel.border = element_blank()
      )
  return(plt)
}
# plot
plt_list_temp <- 1:(layers_grp %>% terra::nlyr()) %>% 
  purrr::map(plt_fn_temp)

# combine
cowplot::plot_grid(
  plotlist = plt_list_temp
  , nrow = 1
)
```

### Map of grouped Canopy Cover and HLI

```{r mapview-grp, results='asis'}

l1_temp <- layers_grp[[1]] %>% 
  stars::st_as_stars()
l2_temp <- layers_grp[[2]] %>% 
  stars::st_as_stars()
mapview::mapviewOptions(
  homebutton = FALSE
  , basemaps = c("Esri.WorldImagery","OpenTopoMap")
  , na.color = "transparent"
)
mapview::mapview(
  l1_temp
  , col.regions = viridis::turbo(n=5)
  , legend = TRUE
  , layer.name = names(layers_grp[[1]])
) + 
mapview::mapview(
  l2_temp
  , col.regions = viridis::turbo(n=5)
  , legend = T
  , layer.name = names(layers_grp[[2]])
  , hide = T
) 

```

### Combine Grouped Canopy Cover and HLI 

```{r class-combine}
# combine layers for classification
  concat_rast <-terra::concats(layers_grp[[1]], layers_grp[[2]]) %>% 
    setNames(
      (layers %>% names() %>% paste(collapse = "_"))
    )
# data_frame
concat_df <- concat_rast %>%
    as.data.frame(xy=T) %>% 
    separate_wider_delim(
      crowncov_hli
      , delim = "_"
      , names=c("crowncov", "hli")
      , cols_remove = F
    ) %>% 
    dplyr::mutate(
      crowncov_hli_filter = forcats::fct_other(
        f = crowncov_hli
        , keep = c(
          "1_1","1_3","1_5"
          , "3_1","3_3","3_5"
          , "5_1","5_3","5_5"
        )
        , other_level = "other"
      )
      , dplyr::across(
          c(crowncov,hli)
          , ~ factor(
              .x
              , levels = 1:5
              , labels = c("low","med.low","medium","med.high","high")
              , ordered = T
            )
        )
      , lat = y
      , lon = x
    ) %>% 
    # points
    sf::st_as_sf(coords=c("lat","lon"),crs=terra::crs(crowns_rast))
# plot
   concat_df %>% 
    dplyr::filter(crowncov_hli_filter!="other") %>% 
    ggplot() + 
      geom_tile(aes(x=x,y=y,fill=crowncov_hli_filter)) +
      geom_sf(data = bhef_boundary, alpha = 0, lwd = 1, color = "black", linetype = "dashed") +
      geom_sf(data = rx_fire, alpha = 0, lwd = 1, color = "orangered") +
      geom_sf_label(data = rx_fire, aes(label = unit)
        , label.size = NA
        , alpha = 0
        , size = 3
      ) +
      coord_sf(expand = F) +
      scale_fill_viridis_d(option = "turbo", alpha = 0.8) + 
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      labs(
          title = "Canopy Cover (1=low,3=med,5=hi)\nx HLI (1=low,3=med,5=hi)"
          , subtitle = paste0("Rx fire units")
          , x=""
          , y=""
          , fill = "Canopy_HLI"
        ) +
      theme_light() +
      theme(
        legend.position = "top"
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        # , legend.box.margin = margin(-21,0,2,0)
        , legend.text = element_text(size = 7)
        , legend.title = element_text(size = 7)
        , axis.text = element_text(size = 7)
        , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
        , plot.title = element_text(size = 10, face = "bold", margin = margin(0,0,2,0))
        , plot.subtitle = element_text(size=8, face="italic", color="orangered", margin = margin(0,0,0,0))
        , panel.border = element_blank()
      ) + 
      guides(fill=guide_legend(nrow=3,byrow=TRUE))
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()

concat_df %>% str()

ggplot(
  data = concat_df %>% 
    sf::st_filter(
      bhef_boundary %>% 
        sf::st_centroid() %>% 
        sf::st_buffer(20)
    )
) + 
  geom_sf(mapping = aes(color = crowncov_hli_filter), shape = ".")
```
